---
title: "Early gigantic lamniform marks the onset of large-body size in modern shark evolution"
title-block-banner: "#2E2D29"
title-block-banner-color: "#FFFFFF"
subtitle: "Supplementary Data:"
description: "Stanford Doer School of Sustainability, Paleobiology Lab"
date: last-modified
date-format: "YYYY-MM-DD"
mainfont: Figtree
sansfont: Georgia
footnotes-hover: true
reference-location: margin
lang: en
search: true
number-sections: false
crossref:
  chapters: true
author:
  - name: Mohamad Bazzi
    orcid: 0000-0002-9495-0781
    email: bazzi@stanford.edu
    url: https://github.com/mohabazzi
    affiliations:
      - name: Stanford University
        department: Department of Earth and Planetary Sciences
        address: CA 94305
        country: USA
        url: https://sustainability.stanford.edu/
highlight-style: pygments
fig-cap-location: top
lightbox: auto
format:
  stanford-quarto-html: default
other-links:
  - text: Custom Quarto Template
    icon: file-code
    href: https://github.com/mohabazzi/stanford-quarto
  - text: Docker image
    icon: file-code
    href: https://hub.docker.com/r/mohamadbazzi/aptian-image
code-links:
  - text: Github Repo
    icon: github
    href: https://github.com/mohabazzi/AptianShark
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

```{css}
.sidebar nav {
  position: sticky;
  top: 80px;
}
```

## Article summary

::: {.callout-summary}

Lamniform sharks include some of the largest-bodied organisms on Earth and trace their origin back ~135 million years. A growing, although exiguous archive of fossilized remains from the nascent stages of lamniform shark evolution indicates that large body sizes evolved long before lamniform species replaced marine tetrapods as top predators during the early half of the Late Cretaceous. The evolution of large body size in sharks has been linked with both physiological and environmental adaptations, in particular endothermy, and the ecological expansion into oceanic waters. Here, we report the oldest known giant lamniform shark, diagnosed from isolated vertebral centra and discovered in a ~115-million-year-old inner-neritic deposit (Aptian, Early Cretaceous) from Darwin, Northern Territory, Australia. Using a comprehensive dataset of living lamniform sharks comprising 1,912 individuals, we built and fitted both intra-and interspecific regression models based on the relationship between vertebral centrum diameter and total length. These models were applied to predict the body size of this new giant shark alongside other Cretaceous lamniform species known from vertebral elements. Our main results demonstrate that giant body sizes (≥ 6 meters) in lamniform sharks have an ancient origin, with our new fossil discovery supporting a neritic hypothesis for the evolutionary emergence of such phenotypic trait. Furthermore, notwithstanding the challenge in reconstructing the trophic ecology of extinct species, our new size estimates underscore a continuity and diversity of large body sizes throughout lamniform evolutionary history.

:::

## Libraries

```{r message=FALSE,results='hide',warning=FALSE}
R.packages <- c("ozmaps ggpubr ggspatial dplyr openxlsx ggplot2 tidyr reshape janitor downlit
                 purrr readxl kableExtra knitr gtsummary gt MASS broom emmeans geomorph xml2
                 smatr MASSTIMATE ggpmisc ggrepel smatr performance flextable Metrics
                 rempsyc car forcats RRPP caper geiger phangorn lmtest randomForest
                 MuMIn phytools conflicted tibble irr nlme ggdist adephylo")

import_packages <- 
  function(x)
  x |> trimws() |> strsplit("\\s+") |> unlist() |> 
  lapply(function(x) library(x, character.only = T)) |> 
  suppressMessages() |> 
  suppressWarnings()

R.packages |> import_packages()

# Resolve conflicted functions.
conflicted::conflict_prefer("map", "purrr")
conflicted::conflict_prefer("filter", "dplyr")
conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("rename","dplyr")
conflicted::conflict_prefer("summarize", "dplyr")
conflicted::conflict_prefer("annotate", "ggplot2")
conflicted::conflict_prefer("AICc", "MuMIn")
```

## Custom functions

```{r}
#' @title Compute average measurements across multiple dataframes
#' 
#' @description #' This function takes a list of data frames and a set of variable names, and returns a single data frame containing the average values of those 
#' variables across all input data frames.
#' 
#' @return A dataframe.
#' @details Remove is given as an argument to filter the 'replicate column' which is numeric and losses meaning after Reduce is applied.

aggregate_measurements <- function(path = NULL, remove = NULL) { 
  sheets <- getSheetNames(file = path)
  numeric_var <- lapply(sheets, function(x) {
    df <- read.xlsx(xlsxFile = path, sheet = x)
    df |> select(where(is.numeric))
  }) 
  names(numeric_var) <- sheets
  # Compute the average.
  ave_sum <- Reduce(f = "+",x = numeric_var)/length(numeric_var)
  if (!is.null(remove)) {
    ave_sum <- ave_sum |> select(-all_of(remove))
  }
  return(ave_sum)
}

other_details <- function(path = NULL) {
  frame <- 
    read.xlsx(xlsxFile = path, sheet = 1) |> 
    # Bring back species and sex information. 
    select(where(~ !is.numeric(.)))
    return(frame)
}

#' @title Intraclass correlation coefficient
#' @description 
#' @return 
#' @details 

icc_custom <- function(path = NULL) {
  sheet_names <- getSheetNames(path)
  # List.
  sheet_lists <- 
    map(sheet_names, ~ read.xlsx(path, sheet = .x))
  # Select variable.
  sheet_lists <- 
    map(sheet_lists, ~ .x |> select("centrum_diameter_cm","total_length_cm")) |> 
    lapply(as.data.frame)
  # Combine into master object.
  combined <- do.call(cbind,sheet_lists)
  
  names(combined) <- make.unique(names(combined))
  
  centrum <- combined |> select(matches("^centrum"))
  length  <- combined |> select(matches("length"))
  
  ls_obj  <- list(centrum,length)
  names(ls_obj) <- c("Centrum","Length")
  # Compute ICC.
  irr_centrum <- irr::icc(ratings = ls_obj$Centrum)
  irr_length  <- irr::icc(ratings = ls_obj$Length)
  
  irr_centrum_res <- 
    cbind(irr_centrum$subjects,irr_centrum$raters,
          irr_centrum$value,irr_centrum$r0,
          irr_centrum$p.value) |>
  as.data.frame() |> 
  setNames(nm = c("subjects","raters","value","r0","p-value"))
  
  irr_length_res <- 
    cbind(irr_length$subjects,irr_length$raters,
          irr_length$value,irr_length$r0,
          irr_length$p.value) |>
  round(digits = 1) |> 
  as.data.frame() |> 
  setNames(nm = c("subjects","raters","value","r0","p-value"))
  
  # Combine results.
  combin_res <- 
    bind_rows(irr_centrum_res,irr_length_res) |> 
    mutate(variable = c("Centrum","Total Length"))
  
  return(combin_res)
}

#'@title Confidence and Prediction Intervals.

intervals <-  function(xvar,alpha=0.975) {
  mean<-mean(xvar)
  std<-sd(xvar)
  n<-length(xvar)
  ci.s.x<-sqrt((std^2)/n)
  pi.s.x<-std*sqrt(1+1/n)
  t<-qt(alpha,df=n-1)
  ci.uci<-mean+t*ci.s.x
  ci.lci<-mean-t*ci.s.x
  pi.uci<-mean+t*pi.s.x
  pi.lci<-mean-t*pi.s.x
  res<-as.list(c(round(mean,4),round(std,4),n,round(t,4),round(ci.lci,4),round(ci.uci,4),round(pi.lci,4),round(pi.uci,4)))
  names(res)<-c('mean','st.dev','N','t.stat','Lower.CI','Upper.CI','Lower.PI','Upper.PI')
  return(res)
}
```

## Aims

Here we adopt an extant-scaling approach using Ordinary Least Squares regression (OLS) modelling based on the total length (TL) and centrum diameter (CD) of modern lamniform species to **1)** explore the performance of centrum diameter as a proxy of body size; and **2)** generate body size predictions (total length and mass) in the hitherto largest described and geologically oldest giant lamniform shark on record.

## Fossil site

```{r warning=FALSE,message=FALSE,results='hide',echo=FALSE,fig.show='hide'}
# Locality map.
AU <- ozmaps::ozmap("states")
```

```{r warning=FALSE, message=FALSE,fig.cap="Geographic map of Casuarina Beach, Darwin, Northern Territory, Australia. The fossil locality is indicated by a red rectangle. The maps were generated programmatically using the ozmap and ggplot2 packages."}
AuMap <- 
  ggplot(data = AU) +
  geom_sf(fill= "antiquewhite") +
  annotate(geom = "rect",xmin = 130.5, xmax = 131.2, ymin = -12.7, ymax = -12.2, color = "#E16A86", fill = NA, lwd = .8) +
  xlab("Longitude") + 
  ylab("Latitude") +
  theme_bw() + 
  theme(panel.grid.major = element_line(color = gray(.5), linetype = "dashed", linewidth = 0.5), 
        panel.background = element_rect(fill = "aliceblue"))

# Casuarina Beach, Darwin, Northern Territory.
ggarrange(AuMap,
          AU |> 
  filter(NAME == "Northern Territory") |> 
  ggplot(aes(fill = NAME)) + 
  geom_sf(fill = "antiquewhite",show.legend = F) +
  labs(x= "Longitude", y = "Latitude") +
  # Darwin.
  geom_rect(aes(xmin = 130.5, xmax = 131.2, ymin = -12.7, ymax = -12.2), color = "#E16A86", fill = NA, lwd = .8) +
  # Casuarina Beach
  geom_point(aes(x = 130.873794555, y = -12.352499961), color = "black", size = 3,show.legend = F) +
  geom_label(aes(x = 132, y = -13, label = "Casuarina Beach, Darwin"),color = "black", fontface = 2, fill = NA) +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true", pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"), 
                         style = north_arrow_fancy_orienteering) +
  theme_bw() + 
  theme(plot.title = element_text(face = "bold"),
        panel.grid.major = element_line(color = gray(.5), linetype = "dashed", linewidth = 0.5),
        panel.background = element_rect(fill = "aliceblue")))
```

## Data sources {.smaller}

### Data wrangling and feature engineering

**Extant size data:** **\[1\]** Data on body length and mass, along with measurements of vertebral centra of white sharks (*Carcharodon carcharias*) off the east coast of South Africa [@Winter1999]. **\[2\]** Equivalent measurement data compiled from the literature of Lamnidae and Alopiidae species: *Isurus oxyrinchus*, *Lamna nasus*, and *Alopias pelagicus* [@Natanson2006; @Natanson2002; @Drew2015]. **\[3\]** Complementary data extracted from static images of scatter plots showing the relationship between centrum diameter/radius and body length (fork, precaudal, and total length) of *Pseudocarcharias kamoharai* [@Kindong2020], *Cetorhinus maximus* [@Natanson2008], *Lamna ditropis* [@Goldman2006], *Carcharias taurus* [@Lucifora2003], *A. pelagicus* [@Liu1999], *C. carcharias* [@Cailliet1985], *A. vulpinus* [@Cailliet1983], and *A. superciliosus* [@Liu1998]. **\[4\]** Single specimen measurements of *Odontaspis noronhai* [@Stone2019] and *O. ferox* [@Hansen2013]. **\[5\]** Vertebral centrum measurements, body size, sex, and vertebral count along the spinal column of captured *C. carcharias*, *L. nasus*, *I. oxyrinchus*, and *A. vulpinus* [@Natanson2018].

**Fossil size data:** Literature-sourced data on body length and maximum vertebral centra from multiple Cretaceous lamniform species.

**Input data:** Five Aptian-aged cardabiodontid shark vertebral centra of unknown body size.

```{r}
# Alopias pelagicus (references: Liu et al. 1999, fig.1 and Drew et al. 2015).
alopias_pelagicus_liu <- 
  cbind(
    aggregate_measurements(path = "data/Liu et al. 1999.xlsx",remove = "replicate"),
    other_details(path = "data/Liu et al. 1999.xlsx")
  ) |>
  relocate("species",.before = 1)

alopias_pelagicus_drew <- 
  cbind(
    aggregate_measurements(path = "data/Drew et al. 2015.xlsx"),
    other_details(path = "data/Drew et al. 2015.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1)

# Combine the two sources.
alopias_pelagicus_data <- 
  bind_rows(
    alopias_pelagicus_liu |>
      select(any_of(c("order","family","genus","species","centrum_diameter_cm","total_length_cm","reference"))),
    alopias_pelagicus_drew |> 
      select(any_of(c("order","family","genus","species","centrum_diameter_cm","total_length_cm","reference")))
  )

# Alopias superciliosus (reference: Liu et al. 1998).
alopias_superciliosus_data <- 
  cbind(
  aggregate_measurements(path = "data/Liu et al. 1998.xlsx",remove = "replicate"),
  other_details(path = "data/Liu et al. 1998.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1) |> 
  select(!any_of(c("centrum_radius_mm","centrum_diameter_mm","specimen","total_length_mm","precaudal_length_cm","sex")))

# Alopias vulpinus (reference: Cailliet et al. 1983)
alopias_vulpinus_data <- 
  cbind(
    aggregate_measurements(path = "data/Cailliet et al. 1983.xlsx",remove = "replicate"),
    other_details(path = "data/Cailliet et al. 1983.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1) |> 
  select(!any_of(c("centrum_diameter_mm","specimen","total_length_mm")))

# Carcharias taurus (reference: Lucifora 2003, fig. 6.2)
carcharias_taurus_data <- 
  cbind(
    aggregate_measurements(path = "data/Lucifora 2003.xlsx",remove = "replicate"),
    other_details(path = "data/Lucifora 2003.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1) |> 
  select(!any_of(c("centrum_radius_mm","centrum_diameter_mm","specimen")))

# Carcharodon carcharias (reference: Cailliet et al. 1985, fig. 3, and Wintner and Cliff 1999, fig. 3)
carcharodon_carcharias_cailliet <-
  cbind(
    aggregate_measurements(path = "data/Cailliet et al. 1985.xlsx",remove = "replicate"),
    other_details(path = "data/Cailliet et al. 1985.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1) |> 
  select(!any_of(c("centrum_radius_mm","centrum_diameter_mm","total_length_mm")))

carcharodon_carcharias_wintner <-
  cbind(
    aggregate_measurements(path = "data/Wintner & Cliff 1999.xlsx"),
    other_details(path = "data/Wintner & Cliff 1999.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1) |> 
  select(!any_of(c("LAB","CAPDATE","CDPRED","TL_Measured"))) |> 
  rename("precaudal_length_cm" = "PCL(cm)",
         "centrum_diameter_cm" = "CD(cm)",
         "Tagno" = "TAGNO",
         "netmass_kg" = "NETMASS",
         "sex" = "Sex") |> 
  mutate(sex = ifelse(sex == 1,"male","female")) |> 
  mutate_if(is.character,as.factor) |> 
  mutate(sex = relevel(sex,"male")) |> 
  # Converting-factor to TL (Total Length)
  mutate(total_length_cm = 1.251 * precaudal_length_cm + 5.207) |> 
  mutate(log_tl = log10(total_length_cm),
         log_cd = log10(centrum_diameter_cm),
         log_netmass = log10(netmass_kg)) |> 
  # Reshuffle column order.
  relocate("total_length_cm",.after = "netmass_kg") |> 
  # Remove row with missing values in key variables.
  filter(!is.na(netmass_kg))

# Cetorhinus maximus (reference: Natanson et al. 2008, sample size = 21)
cetorhinus_maximus_data <- 
  cbind(
    aggregate_measurements(path = "data/Natanson et al. 2008.xlsx",remove = "replicate"),
    other_details(path = "data/Natanson et al. 2008.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1)

# Lamna nasus (reference: Natanson et al. 2002, sample size = 575)
lamna_nasus_data <- 
  cbind(
    aggregate_measurements(path = "data/Natanson et al. 2002.xlsx"),
    other_details(path = "data/Natanson et al. 2002.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1) |>
  relocate("centrum_diameter_cm",.before = "total_length_cm") |> 
  select(!any_of(c("sample_number","number_of_band","centrum_radius_mm","fork_length_cm","centrum_diameter_mm","sex")))

# Lamna dipturs (reference: Goldman & Musick 2006, fig.2)
lamna_dipturs_data <- 
  cbind(
    aggregate_measurements(path = "data/Goldman & Musick 2006.xlsx",remove = "replicate"),
    other_details(path = "data/Goldman & Musick 2006.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1) |> 
  select(!any_of(c("centrum_radius_mm","centrum_diameter_mm","precaudal_length_cm")))

# Isurus oxyrinchus (reference:Natanson et al. 2006)
isurus_oxyrinchus_data <- 
  cbind(
    aggregate_measurements(path = "data/Natanson et al. 2006.xlsx"),
    other_details(path = "data/Natanson et al. 2006.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1) |>
  relocate("centrum_diameter_cm",.before = "total_length_cm") |> 
  select(!any_of(c("A","vetebral_radius_mm","fork_length_cm","centrum_diameter_mm","sex")))

# Pseudocarcharias kamoharai (reference: Kindong et al. 2020)
pseudocarcharias_kamoharai_data <- 
  cbind(
    aggregate_measurements(path = "data/Kindong et al. 2020.xlsx",remove = "replicate"),
    other_details(path = "data/Kindong et al. 2020.xlsx")
  ) |> 
  relocate(c("order","family","genus","species"),.before = 1) |> 
  select(!any_of(c("centrum_radius_mm","fork_length_cm","centrum_diameter_mm")))
```

<!-- Group species dataframes-->

```{r}
# Lamniformes.
lamniformes_data <- 
  bind_rows(
  alopias_pelagicus_data,
  alopias_superciliosus_data,
  alopias_vulpinus_data,
  carcharias_taurus_data,
  carcharodon_carcharias_cailliet,
  carcharodon_carcharias_wintner |> select("order","family","genus","species","centrum_diameter_cm","total_length_cm","reference"),
  cetorhinus_maximus_data,
  lamna_nasus_data,
  lamna_dipturs_data,
  isurus_oxyrinchus_data,
  pseudocarcharias_kamoharai_data
) |> 
  mutate(log_tl = log10(total_length_cm),
         log_cd = log10(centrum_diameter_cm)
         ) |> 
  mutate(species = as.factor(species))

# Alopiidae species measurement data.
alopiidae_data <- 
  bind_rows(
  alopias_pelagicus_data,
  alopias_superciliosus_data,
  alopias_vulpinus_data) |> 
  # Log-10 transform variables.
  mutate(log_cd  = log10(centrum_diameter_cm),log_tl = log10(total_length_cm)) |> 
  relocate("reference",.after = 9) # N = 622.

# Combine all extant Lamnidae species.
lamnidae_data <-
  lamniformes_data |> 
  filter(family == "Lamnidae")

# Complete White shark dataset.
carcharodon_carcharias_complete <- 
  lamnidae_data |> 
  filter(species == "Carcharodon carcharias")
```

<!--Auxiliary data -->

```{r message=FALSE}
# Stem-lamniform dataset.
mesozoic_data <- openxlsx::read.xlsx(xlsxFile = "data/mesozoic_data.xlsx",sheet = 1)

# Natanson et al. 2018: white shark data.
Natanson <- "data/Natanson et al. 2018.xlsx"

natanson_data <- 
  excel_sheets(path = Natanson) |>
  set_names() |>
  map(read_excel, path = Natanson)

# Complementary data from Natanson et al. 2018 (non-white shark data)
natanson_supplementary <- read.xlsx(xlsxFile = "data/Natanson et al. 2018_Supplementary.xlsx")
```

<!--Summary tables-->

```{r}
# Table 1.
lamniformes_data |>
  group_by(species) |> 
  # Compute TL range per species.
  summarise("Size Range (cm)" = paste0(round(min(total_length_cm),0), "-", round(max(total_length_cm),0)),
            "Reference" = dplyr::first(reference),
            "N" = n()) |> 
  rename("Species" = "species") |> 
  kable(caption = "Extant taxon sampling. Measurements from 10 lamniform species were sourced. Size-ranges are based on the total length
                   of each species sample (not to be confused with true minimum and maximum body sizes).") |>
  column_spec(column = 1,italic = T) |>
  kable_styling(bootstrap_options = "striped",
                            full_width = F)

# Table 2.
mesozoic_data |>
  relocate(c(species,label), .before = centrum_diameter) |> 
  relocate(c(reference),.after = estimated) |> 
  kable(caption = "Fossil taxon sampling. Measurement data compiled from the literature of 15 lamniform shark species from the Cretaceous period.") |>
  column_spec(column = 1,italic = c(rep(TRUE,3),FALSE,rep(TRUE,2),FALSE,rep(TRUE,8))) |>
  kable_styling(bootstrap_options = "striped",
                            full_width = F)
```

{{< pagebreak >}}

<br/>

### ME Error

```{r}
bind_rows(
    icc_custom(path = "data/Liu et al. 1999.xlsx") |> mutate(species = "Alopias pelagicus"),
    icc_custom(path = "data/Liu et al. 1998.xlsx") |> mutate(species = "Alopias superciliosus"),
    icc_custom(path = "data/Cailliet et al. 1983.xlsx") |> mutate(species = "Alopias vulpinus"),
    icc_custom(path = "data/Lucifora 2003.xlsx") |> mutate(species = "Carcharias taurus"),
    icc_custom(path = "data/Cailliet et al. 1985.xlsx") |> mutate(species = "Carcharodon carcharias"),
    icc_custom(path = "data/Natanson et al. 2008.xlsx") |> mutate(species = "Cetorhinus maximus"),
    icc_custom(path = "data/Goldman & Musick 2006.xlsx") |> mutate(species = "Lamna ditropis"),
    icc_custom(path = "data/Kindong et al. 2020.xlsx") |> mutate(species = "Pseudocarcharias kamoharai")
  ) |> 
  mutate_at(.vars = "p-value",.funs = ~round(.x,digits = 3)) |> 
  mutate(interpretation = case_when(value >= 0.9 ~ "Excellent agreement",TRUE ~ "")) |> 
  kable() |> 
  kable_styling(bootstrap_options = "striped",full_width = F)
```

## Linear regression

This section explores the relationship between TL-CD in extant male and female *C. carcharias* using OLS-regression modelling. A factorial anova was also used to test the influence of centrum diameter and sex on total length. Table results show that both `log_cd` and `sex` have significant effects on the dependent variable `log_tl` while their interaction does not play a significant role in *C. carcharias*. These results are consistent with those presented in the original account by [@Winter1999].

::: rmdnote
$$log10(y) = \log10(\alpha) + log10(\beta) \times log10(x) + \epsilon$$

**Where:**

$y$ represents the dependent variable $x$ represents the independent variable $\beta_0$ is the intercept or the value of $y$ when $x$ equals 0 $\beta_1$ is the coefficient or the amount by which $y$ changes when $x$ increases by one unit $\epsilon$ represents the error term or the variability in $y$ that cannot be explained by $x$.
:::

```{r}
# Linear model specification.
MEq <- summary(lm(log_tl~log_cd, data = carcharodon_carcharias_wintner,subset = carcharodon_carcharias_wintner$sex == "male"))
FEq <- summary(lm(log_tl~log_cd, data = carcharodon_carcharias_wintner,subset = carcharodon_carcharias_wintner$sex == "female"))

# Coefficients.
MCe <- unlist(as.character(round(MEq$coefficients[,1],4)))
MCe <- "Y_male = 1.9238 + 0.8873 * x"
FCe <- unlist(as.character(round(FEq$coefficients[,1],4)))
FCe <- "Y_female = 1.9444 + 0.8704 * x"

# Factorial anova test.
aov(formula = log_tl~log_cd*sex,data = carcharodon_carcharias_wintner) |>
  tidy() |> 
  setNames(nm = c("Term","Df","SS","MSS","F-statistic","P-value")) |> 
  knitr::kable(caption = "Factorial anova test") |>
  kable_styling(bootstrap_options = "striped",full_width = F)
```

<br/>

### Model fit I (TL)

See section @sec-gws for regression results based on all *C. carcharias* data.

```{r message=FALSE, results='hide'}
# Linear model specification (OLS: First degree)
length_linear_fit_ols <- lm(log_tl~log_cd, data = carcharodon_carcharias_wintner)

# Summary.
# length_linear_fit_ols |> summary() |> broom::tidy(conf.int = TRUE)
tbl_regression(length_linear_fit_ols, include = everything(),intercept = TRUE) |>
  as_gt() |>  
  tab_header("Linear regression model output (CD ~ TL)") |>
  tab_options(table.align='left')

# Model Accuracy.
round(broom::glance(length_linear_fit_ols)$sigma,digits = 3) # RSE.
# Model Error (Residual sum of squares, RSS = deviance)
# Residuals ( = the lack of fit)
length_linear_fit_ols |> 
  resid() |>
  crossprod() |> 
  cat() # RSS

# Compare with Robust linear regression.
length_linear_fit_rlm <- rlm(log_tl~log_cd, data = carcharodon_carcharias_wintner)
summary(length_linear_fit_rlm)$sigma # Residual standard error (no major difference from the OLS fit)

# Compare models.
AICc(length_linear_fit_ols,length_linear_fit_rlm)

# Test if slope is equal to 1 (i.e. isometry).
isomodel_01 <- 
  smatr::slope.test(y = carcharodon_carcharias_wintner$log_tl,
                    x = carcharodon_carcharias_wintner$log_cd,
                    test.value = 1,method = "OLS") # Negative allometry.

# Mean Percent Prediction Error (PPE).
ppe_01 <-
  ppe(true = 10^carcharodon_carcharias_wintner$log_tl,
      pred = 10^as.vector(fitted(length_linear_fit_ols)))
```

```{r warning=FALSE, message=FALSE, fig.cap="Relationship between centrum diameter and total length in C. carcharias (male and female). Observations (i.e., individuals) from the original dataset by Wintner & Cliff (1999) with missing mass estimates were dropped (n = 3). Fitted model equation derived from OLS."}
# Plot.
carcharodon_carcharias_wintner |> 
  ggplot(mapping = aes(x = log_cd,y = log_tl),group = sex, colour = sex, fill = sex) +
  annotation_logticks() + 
  geom_point(aes(color = sex,size = log_netmass),
             shape = 19) +
  scale_color_manual(values = c("#B83A4B","#016895")) +
  scale_fill_manual(values =  c("#B83A4B","#016895")) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  ggpmisc::stat_poly_eq(use_label(c("eq", "adj.R2", "f", "p", "n")),
                        size = 4.0,label.x = "left",label.y = "top",
                        color = "black") +
  # RLM line.
  geom_abline(slope = 0.8728,intercept = 1.9385,colour = "#016895",linetype = 2) +
  # Labels.
  labs(size = "Body mass (log10 kg)",x = "Centrum diameter (log10 cm)", y = "Total length (log10 cm)",
       colour = "sex",
       title = NULL,
       subtitle = NULL,
       caption = "") +
  geom_label_repel(aes(label = Tagno),
                  box.padding   = 0.35, 
                  point.padding = 0.5,max.overlaps = 10,
                  segment.color = 'grey50') +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)
```

### Regression diagnostics (TL)

```{r fig.cap="Visual check of model assumptions (i.e. Linearity, Independence, Normality and Equal Variances)."}
# Linear model assumption.
performance::check_normality(length_linear_fit_ols)
performance::check_model(x = length_linear_fit_ols,
                         check = c("normality", "linearity","homogeneity", "outliers"),
                         axis_title_size = 8)
```

```{r}
# Assumptions checks.

# kable(nice_assumptions(model = length_linear_fit_ols), col.format.p = 2:4,
#       caption = "Diagnostics for the linear model (CD-TL) applied to C. carcharias.
#                  Diagnostic is how many assumptions (i.e. normality of residuals, homoscedasticity, and independence) are not respected by the model fit.
#                  P-values < .05 imply assumptions are not respected.") |>
#   kable_styling(bootstrap_options = "striped",full_width = T)
```

<br/>

### Model fit II (BM)

```{r warning=FALSE, message=FALSE, results='hide'}
# Fit body mass linear model with OLS.
mass_linear_fit_ols <- lm(log_netmass ~ log_cd, data = carcharodon_carcharias_wintner)
# Summary.
mass_linear_fit_ols |> 
  summary() |> 
  broom::tidy(conf.int = TRUE)

# Accuracy.
broom::glance(mass_linear_fit_ols)

# RLM.
mass_linear_fit_rlm <- rlm(log_netmass~log_cd, data = carcharodon_carcharias_wintner)
# RSE.
summary(mass_linear_fit_rlm)$sigma

# Test if slope is equal to 3.
isomodel_02 <- smatr::slope.test(y = carcharodon_carcharias_wintner$log_netmass,
                                 x = carcharodon_carcharias_wintner$log_cd,
                                 test.value = 3,method = "OLS") # negative allometry.

# lrtest(mass_linear_fit_ols,mass_linear_fit_rlm)

# Mean Percentage Prediction Error (PPE).
ppe_02 <-
  ppe(true = 10^carcharodon_carcharias_wintner$log_netmass,
      pred = 10^as.vector(fitted(mass_linear_fit_ols)))
```

```{r warning=FALSE,message=F, fig.cap="Relationship between the logarithm of centrum diameter and mass (A) and centrum diameter and mass on the measurement scale (B) in C. carcharias."}

# Prediction intervals.
piBM <- 
  predict(object = mass_linear_fit_ols,newdata = data.frame(log_cd = carcharodon_carcharias_wintner$log_cd),interval = "prediction") |> 
  as.data.frame()

ggarrange(
  # 1. CD vs Mass on log-scales.
  carcharodon_carcharias_wintner |> 
  ggplot(mapping = aes(x = log_cd,y = log_netmass),group = sex, colour = sex, fill = sex) +
  annotation_logticks() + 
  geom_point(aes(color = sex),shape = 19) +
  scale_color_manual(values = c("#B83A4B","#016895")) +
  scale_fill_manual(values =  c("#B83A4B","#016895")) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  # Prediction intervals.
  geom_line(aes(y = piBM$lwr), col = "black", linetype = "dashed") +
  geom_line(aes(y = piBM$upr), col = "black", linetype = "dashed") +
  ggpmisc::stat_poly_eq(use_label(c("eq", "adj.R2", "f", "p", "n")),
                        size = 4.0,label.x = "left",label.y = "top",
                        color = "black") +
  # Labels.
  labs(x = "Centrum diameter (log10 cm)", y = "Body mass (log10 kg)",
       colour = "Sex") +
  geom_label_repel(aes(label = Tagno),
                  box.padding   = 0.35, 
                  point.padding = 0.5,max.overlaps = 10,
                  segment.color = 'grey50') +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "none",
        legend.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1),
  # 2. CD vs Mass on raw measurment scale.
  carcharodon_carcharias_wintner |> 
  ggplot(mapping = aes(x = centrum_diameter_cm,y = netmass_kg),group = sex, colour = sex, fill = sex) +
  geom_point(aes(color = sex),shape = 19) +
  scale_color_manual(values = c("#B83A4B","#016895")) +
  scale_fill_manual(values =  c("#B83A4B","#016895")) +
  stat_poly_line(aes(colour = sex),formula = y ~ poly(x, 3, raw = TRUE)) +
  stat_poly_eq(aes(colour = sex, label = after_stat(eq.label)),formula = y ~ poly(x, 3, raw = TRUE)) +
  # Labels.
  labs(x = "Centrum diameter (cm)", y = "Body mass (kg)",colour = "Sex") +
  geom_label_repel(aes(label = Tagno),
                  box.padding   = 0.35, 
                  point.padding = 0.5,max.overlaps = 10,
                  segment.color = 'grey50') +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "none",
        legend.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1),
  # Add legend.
  common.legend = F,align = "hv")
```

### Leverage analysis (BM)

Here we explore the presence of unusual and influential observations in the BM model.

<!--Regression diagnostics-->

```{r message=FALSE, results='hide',fig.cap="Regression diagnostics."}
# Regression diagnostics.
performance::check_normality(mass_linear_fit_ols)
performance::check_model(x = mass_linear_fit_ols,
                         check = c("normality", "linearity","homogeneity", "outliers"))
```

```{r}
# One assumption (i.e. normality) is not respected.

# kable(nice_assumptions(model = mass_linear_fit_ols), stars = T, col.format.p = 2:4) |>
#   kable_styling(bootstrap_options = "striped",
#                 full_width = T)
```

<!--Leverage analysis-->

```{r fig.cap="Leverage analysis. The size of the bubbles in the plot indicates the Cook’s D value of each data point."}
cook.D <- cooks.distance(model = mass_linear_fit_ols)
# Plot.
layout(matrix(c(1,2,3), nrow = 1, ncol = 3,byrow = FALSE),
       respect = TRUE,widths = c(4,4,4),heights = c(6,6,6))
# Q-Q plot.
invisible(capture.output(
  car::qqPlot(mass_linear_fit_ols, main="Q-Q plot", distribution="norm",ylab = "Studentized Residuals",
              col = "#B83A4B", 
              col.lines=carPalette()[1],
              pch = 19)
)
)

# Cooks distance.
plot(cook.D, pch=21, cex=1, main="Influential observations by Cooks distance", 
     col = "black",bg = "white",font.lab = 2, cex.main = 1,bty = "n")

abline(h = 4/nrow(carcharodon_carcharias_wintner), col="#B83A4B",lwd = 2)
# Regression Influence Plot.
invisible(capture.output(car::influencePlot(model = mass_linear_fit_ols, fill.col=carPalette()[5],font.lab = 2)))
```

<!--Nested model comparison-->

```{r results='hide'}
# Model improvement.
mass_linear_fit_ols_2 <- lm(log_netmass ~ log_cd, data = carcharodon_carcharias_wintner[cook.D < 4/length(cook.D),])
#summary(mass_linear_fit_ols_2)

# Residual standard error (RSE) and R-squared.
c(summary(mass_linear_fit_ols)$sigm,summary(mass_linear_fit_ols_2)$sigma
) |> setNames(nm = c("Model I","Model II"))

c(summary(mass_linear_fit_ols)$r.squared,summary(mass_linear_fit_ols_2)$r.squared
) |> setNames(nm = c("Model I","Model II"))

# Check diagnostics.
# nice_table(nice_assumptions(model = mass_linear_fit_ols_2),stars = T, col.format.p = 2:4)

# Quadratic model (or second-degree polynomial).
mass_quad_fit_ols <- lm(log_netmass~log_cd+I(log_cd^2),data = carcharodon_carcharias_wintner,subset = cook.D < 4 / length(cook.D))
mass_quad_fit_ols |> 
  summary()

# Model selection.
lrtest(mass_linear_fit_ols_2,mass_quad_fit_ols)  # No improvement.
AICc(mass_linear_fit_ols_2,mass_quad_fit_ols)

# Mean Percentage Prediction Error (PPE).
ppe_03 <-
  ppe(true = 10^carcharodon_carcharias_wintner[cook.D < 4/length(cook.D),]$log_netmass,
      pred = 10^as.vector(fitted(mass_linear_fit_ols_2)))
```

### Fossil Predictions (I)

#### Total length

In this section, **TL and BM** estimates are based on the white shark data from [@Winter1999]. Point estimates, 95% prediction and confidence intervals are computed for each specimen. 

Prediction are given by:

$$Y_i = \hat{\alpha} + \hat{\beta} \cdot X_i$$

**Specimen remarks**

P922-33: one side is cracked and artificially ‘swollen’ measuring 129 mm; the other un-deformed side is 125 mm. P922-32: 126 mm but inflated due to cracking. P9382: 118 mm. P922-37: 114 mm but most of the surface covered with nodules so measurement might be artificially inflated. P922-29: 117 mm as preserved but greatest radius is 63 mm which indicates an original 126 mm.

```{r}
# Fossil shark centrum diameter in centimeters.
aptian_shark <- data.frame(log_cd = log10(c(12.5,12.6,11.8,11.4,12.6)))
```

```{r}
# Total length (TL) estimates with 95% CI & PI.
pred_tl <-
  cbind(10^predict(object = length_linear_fit_ols,newdata = aptian_shark,interval = "prediction"),
        10^predict(object = length_linear_fit_ols,newdata = aptian_shark,interval = "confidence")[,2:3]) |> 
  as.data.frame() |> 
  unname() |> 
  setNames(nm = c("fit",".pred_lower",".pred_upper",".conf_lower",".conf_upper")) |>
  # Add specimen and model information.
  mutate(specimen = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29"),
         source = rep("White shark data",5))

# Convert into long format.
long_format <- 
  data.frame(fit = unlist(as.vector(pred_tl$fit)),
             xmin = c(pred_tl$.pred_lower,pred_tl$.conf_lower),
             xmax = c(pred_tl$.pred_upper,pred_tl$.conf_upper),
             intervals = c(rep("prediction",5),rep("confidence",5)),
             specimen = rep(c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29"),2)) |> 
  mutate(specimen = fct_relevel(specimen,c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29")))

# Prediction plot.
tl_graph <-
  long_format |> 
  ggplot(aes(y = specimen, x = fit, colour = intervals,linetype = intervals)) + 
  scale_y_discrete(labels = pred_tl$specimen) +
  geom_errorbar(aes(xmin = xmin, xmax = xmax),
                width = 0.2,linewidth = 1.3,show.legend = T) +
  geom_point(size = 3,fill = "white",
             shape = 21,show.legend = F) +
  # Color intervals.
  scale_fill_manual(values =  c("#4D4D4D","#4D4D4D")) +
  scale_color_manual(values = c("#4D4D4D","#4D4D4D")) +
  # Average TL length
  geom_vline(aes(xintercept=mean(fit)),linetype=1,col = "#4D4D4D", lwd = 1.5) +
  # Largest documented extant great white shark (Carcharodon carcharias) (6.4 m) [Randall JE 1973].
  geom_vline(aes(xintercept = 640),linetype=1,col = "#B2182B", lwd = 1.5) +
  # Text
  annotate("text",x = 700, y = 3.8,
           label = "upper~size~limit~of~italic(C.~carcharias)",
           parse = TRUE,size = 1.5,
           col = "#B2182B",fontface = "bold") +
  # Curve.
  annotate(geom = "curve", x = 700, y = 3.9, xend = 645, yend = 3.8,
          curvature = .3, arrow = arrow(length = unit(2, "mm"))) +
  theme_minimal() +
  labs(x = "Total length (cm)",y = NULL) +
  theme(legend.position = "right",
        title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "plain"),
        axis.text.y = element_text(face = "bold"))

# Apply the bias correction factor.
# pred_tl$fit * exp(summary(length_linear_fit_ols)$sigma^2/2) # corrected factor applied to predictions.

# intervals(xvar = pred_tl$fit)[c(1,7:8)]
```

#### Body mass

```{r warning=FALSE}
# Fossil mass prediction.
pred_mass <-
  cbind(10^predict(object = mass_linear_fit_ols_2,newdata = aptian_shark,interval = "prediction"),
        10^predict(object = mass_linear_fit_ols_2,newdata = aptian_shark,interval = "confidence")[,2:3]) |> 
  as.data.frame() |> 
  unname() |> 
  setNames(nm = c("fit",".pred_lower",".pred_upper",".conf_lower",".conf_upper")) |>
  # Add specimen and model information.
  mutate(specimen = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29"),
         source = rep("White shark data",5))

long_format_mass <- 
  data.frame(fit = unlist(as.vector(pred_mass$fit)),
             xmin = c(pred_mass$.pred_lower,pred_mass$.conf_lower),
             xmax = c(pred_mass$.pred_upper,pred_mass$.conf_upper),
             intervals = c(rep("prediction",5),rep("confidence",5)),
             specimen = rep(c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29"),2)) |> 
  # slice(1:8) |> 
  mutate(specimen = fct_relevel(specimen,c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29")))

# Plot.
mass_graph <- 
  long_format_mass |> 
  ggplot(aes(y=specimen, x=fit, colour=intervals,linetype = intervals)) + 
  scale_y_discrete(labels = pred_mass$specimen) +
  geom_errorbar(aes(xmin = xmin, xmax = xmax),
                width = 0.2,linewidth = 1.3,show.legend = T) +
  geom_point(size = 3,fill = "white", shape = 21,show.legend = F) +
  scale_fill_manual(values =  c("#4D4D4D","#4D4D4D")) +
  scale_color_manual(values = c("#4D4D4D","#4D4D4D")) +
  theme_minimal() +
  # Average BM
  geom_vline(aes(xintercept=mean(fit)),linetype=1,col = "#4D4D4D", lwd = 1.5) +
  # Maximum Great white shark body-weight: <https://www.fisheries.noaa.gov/species/white-shark>
  geom_vline(xintercept = 2041,lwd = 1.5, linetype=1, color = "#B2182B") +
  annotate("text", x = 2350, y = 3.2, label = "upper~size~limit~of~italic(C.~carcharias)",
           parse = TRUE,col = "#B2182B",size = 1.5) +
  labs(x = "Body mass (kg)",y = NULL) +
  theme(legend.position = "right",
        title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "plain"),
        axis.text.y = element_text(face = "bold"))

# intervals(xvar = pred_mass$fit)[c(1,7:8)]
```

```{r fig.cap = "TL and BM estimates for the Darwin cardabiodontid vertebral centra."}
# Combine length and mass predictions graphs.
ggarrange(tl_graph,mass_graph,common.legend = T) + theme(aspect.ratio = .5)
```

```{r fig.cap = "Log-space of C. carcharias centrum size vs the fossil specimens."}
bind_rows(
  carcharodon_carcharias_wintner |> select(log_cd,species),aptian_shark |> mutate(species = "NTM P")) |> 
  ggplot(mapping = aes(x = log_cd, fill = species)) +
  geom_histogram(bins = 30) +
  scale_fill_manual(values = c("#E16A86","black")) +
  theme_bw() +
  theme(axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold",angle = 90,hjust = 0.5,vjust = .5),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "plain"),
        aspect.ratio = 1) +
  labs(title = "Log-10 space",subtitle = "C. carcharias vs. Darwin cardabiodontid vertebral centra",
       x = "Centrum diameter (log10 cm)", y = "Count", fill = "")
```

<!--Visualize main results: Total Length -->

```{r fig.cap = "Predicted TL-values of the Darwin cardabiodontid specimens plotted within the context of the extant white shark data (Wintner & Cliff 1999)."}
# Prepare data.
log_pred_tl <-
  cbind(predict(object = length_linear_fit_ols,newdata = aptian_shark,interval = "prediction"),
        predict(object = length_linear_fit_ols,newdata = aptian_shark,interval = "confidence")[,2:3]) |> 
  as.data.frame() |> 
  unname() |> 
  setNames(nm = c("fit",".pred_lower",".pred_upper",".conf_lower",".conf_upper")) |> 
  mutate(specimen = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29"))

# PI intervals.
PI <- predict(object = length_linear_fit_ols,
              newdata = data.frame(log_cd = carcharodon_carcharias_wintner$log_cd),
              interval = "prediction",se.fit = T,level = .95)

# Prediction interval data subset for plotting.
tl_intervals <- 
  log_pred_tl |> 
  mutate(log_cd = aptian_shark$log_cd)

p <- 
  carcharodon_carcharias_wintner |> 
  ggplot(aes(x = log_cd, y = log_tl)) +
  geom_point(fill = "white",size = 2, shape = 21) +
  geom_abline(intercept = 1.9393, slope = 0.8706,linewidth = 1,colour = "black") +
  annotation_logticks() +
  expand_limits(x = c(0.3443923,1.1003705), y = c(2.228113,2.937239)) +
  # Add bracketing lines.
  geom_vline(xintercept = 0.3443923,linetype = "dashed") +
  geom_vline(xintercept = 0.8267225,linetype = "dashed") +
  # Add prediction intervals.
  geom_line(aes(x = log_cd, y = PI$fit[,"lwr"]), color = "#E16A86") +
  geom_line(aes(x = log_cd, y = PI$fit[,"upr"]), color = "#E16A86") +
  # Add fossil predictions to base plot.
  geom_point(data = aptian_shark,mapping = aes(x = log_cd, y = log_pred_tl$fit), colour = "black", fill = "white",pch = 22, size = 2) +
  # Interpolation line.
  annotate(geom = "segment", x = 0.3443923, xend = 0.8267225, y = 2.8, yend = 2.8,
           colour = "darkgreen",arrow = arrow(type = "closed",length = unit(2, "mm"),ends = "both",angle = 20)) +
  # Extrapolation line.
  annotate(geom = "segment", x = 0.8267225, xend = 1.1003705, y = 2.3, yend = 2.3,
           colour = "darkgreen",arrow = arrow(type = "closed",length = unit(2, "mm"),ends = "both",angle = 20)) +
  labs(title = NULL,subtitle = NULL,
       y = "Total length (log10 cm)",
       x = "Centrum diameter (log10 cm)") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)

# Add error bars.
p0 <- p + 
  geom_errorbar(data = tl_intervals, aes(x = log_cd,y = fit, ymin = .pred_lower,ymax = .pred_upper),
                width = 0.02,color = "black")
```

<!--Body Mass-->

```{r fig.cap = "Predicted BM-values of the Darwin cardabiodontid specimens plotted within the context of the extant white data (Wintner & Cliff 1999)."}
# Prepare data.
log_pred_mass <-
  cbind(predict(object = mass_linear_fit_ols_2,newdata = aptian_shark,interval = "prediction"),
        predict(object = mass_linear_fit_ols_2,newdata = aptian_shark,interval = "confidence")[,2:3]) |> 
  as.data.frame() |> 
  unname() |> 
  setNames(nm = c("fit",".pred_lower",".pred_upper",".conf_lower",".conf_upper"))
  # Raw-scale.
  # mutate(across(everything(), ~ 10^.)) |> 
  # mutate(across(everything(), ~ round(.x,digits = 2)))

# PI intervals.
mPI <- predict(object = mass_linear_fit_ols_2,
               newdata = data.frame(log_cd = carcharodon_carcharias_wintner[cook.D < 4/length(cook.D),]$log_cd),
               interval = "prediction",se.fit = T,level = .95)

# Prediction interval data subset for plotting.
mass_intervals <- 
  log_pred_mass |> 
  mutate(log_cd = aptian_shark$log_cd)

p1 <- 
  carcharodon_carcharias_wintner |> 
  filter(cook.D < 4 / n()) |> 
  ggplot(aes(x = log_cd, y = log_netmass)) +
  geom_point(fill = "white",size = 2, shape = 21) +
  geom_abline(intercept = 0.7598, slope = 2.5089,linewidth = 1,colour = "black") +
  annotation_logticks() +
  expand_limits(x = c(0.3443923,1.1003705), y = c(1.623249, 3.674706)) +
  # Add bracketing lines.
  geom_vline(xintercept = 0.3443923,linetype = "dashed") +
  geom_vline(xintercept = 0.7867514,linetype = "dashed") +
  # Add prediction intervals.
  geom_line(aes(x = log_cd, y = mPI$fit[,"lwr"]), color = "#E16A86") +
  geom_line(aes(x = log_cd, y = mPI$fit[,"upr"]), color = "#E16A86") +
  # Add fossil predictions to base plot.
  geom_point(data = aptian_shark,mapping = aes(x = log_cd, y = log_pred_mass$fit), colour = "black", fill = "white",pch = 22, size = 2) +
  # Interpolation line.
  annotate(geom = "segment", x = 0.3443923, xend = 0.7867514, y = 3.0, yend = 3.0,
           colour = "darkgreen",arrow = arrow(type = "closed",length = unit(2, "mm"),ends = "both",angle = 20)) +
  # Extrapolation line.
  annotate(geom = "segment", x = 0.7867514, xend = 1.1003705, y = 2.0, yend = 2.0,
           colour = "darkgreen",arrow = arrow(type = "closed",length = unit(2, "mm"),ends = "both",angle = 20)) +
  labs(title = NULL,subtitle = NULL,
       y = "Body mass (log10 kg)",
       x = "Centrum diameter (log10 cm)") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)

# Add error bars.
p01 <- p1 + 
  geom_errorbar(data = mass_intervals, aes(x = log_cd,y = fit, ymin = .pred_lower,ymax = .pred_upper),
                width = 0.02,color = "black")
```

### Model fit III (TL)

#### Species-level analysis

We combine all lamniform species data into one master dataframe (N=1912). *O. ferox* and *O. noronhai* are not considered because of low sample size. A series of species-level OLS models are built and fitted. Slope and intercept values are subsequently compared.

```{r message = FALSE, fig.cap = "Bivariate visualizations (log-scale) showing the relationship between CD-TL in lamniform species."}
# Function.
fit_regression <- function(x) {
  lm(log_tl ~ log_cd, x)
}

# Fit linear models.
slope_intercept <- 
  lamniformes_data |> 
  group_by(species) |> 
  do(model = fit_regression(.)) |>
  tibble() |>
  mutate("β0" = map_dbl(model, ~ coef(.)["(Intercept)"]),
         "β1" = map_dbl(model, ~ coef(.)["log_cd"])) |> 
  mutate("y" = paste(round(β0,digits = 2),"+", round(β1,digits = 2),"x")) |> 
  mutate(round(map_df(model, ~ glance(.)[1]),digits = 2)) |> 
  mutate(map_df(model, ~ glance(.)[12])) |> 
  mutate(map_df(model, ~ tidy(.,conf.int = TRUE)["p.value"][2,])) |> 
  mutate(map_df(model, ~ tidy(.,conf.int = TRUE)["conf.low"][2,])) |> 
  mutate(map_df(model, ~ tidy(.,conf.int = TRUE)["conf.high"][2,])) |> 
  mutate("MAE" = round(map_dbl(model, ~ mean(abs(residuals(.)))),digits = 2)) |>
  # select(-model) |> 
  rename("N" = "nobs")

# Pearson correlations.
lamn_corr <- 
  lamniformes_data  |> 
  group_by(species) |> 
  summarize(correlation = cor(log_cd,log_tl, method = "pearson"))

# Pearson correlations on the log-10 values: Entire dataset.
cor(x = lamniformes_data$log_tl,y = lamniformes_data$log_cd) # cor = 0.79 (no different from the raw values: cor = 0.74)

# Color vector.
col_vec <- c("#264653", "#2A9D8F", "#E9C46A", "#F4A261", "#E76F51",
             "#457B9D", "#1D3557", "#F0A8C0FF", "#6D6875", "#A8DADC")

# CD-TL relationship in lamniform species.
intraspecific_graph <- lamniformes_data |> 
  mutate(ecology = case_when(species == "Cetorhinus maximus" ~ "Microphagous", TRUE ~ "Macrophagous")) |> 
  ggplot(aes(x = log_cd, y = log_tl)) +
  scale_fill_manual(values =  col_vec) +
  scale_color_manual(values = col_vec) +
  labs(title = NULL,subtitle = NULL,
       y = "Total length (log10 cm)",
       x = "Centrum diameter (log10 cm)", colour = "Species") +
  annotation_logticks() +
  geom_smooth(aes(colour = species),method = "lm",se = F) + 
  geom_point(aes(color = species), fill = "white",size = 2, shape = 21) +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "none",
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)

# Slopes only.
estimated_slopes <- lamniformes_data |>
  group_by(species) |>
  do(tidy(lm(log_tl ~ log_cd, data = .),conf.int = T)) |>
  dplyr::select(species,term,estimate,contains("conf")) |>
  rename("Estimate" = "estimate") |>
  filter(term != "(Intercept)") |>
  ggplot(mapping = aes(x = fct_rev(species), y = Estimate)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                color = col_vec,
                position = position_dodge(0.4), width = .3,show.legend = F) +
  geom_point(aes(color = species),shape = 21,fill = "white",show.legend = F) +
  geom_hline(yintercept = c(1), linetype = "dashed") +
  scale_fill_manual(values = col_vec) +
  scale_color_manual(values = col_vec) +
  labs(x = NULL) +
  coord_flip() +
  theme_bw() +
  theme(aspect.ratio = 1,
        panel.grid.major.y = element_blank(),
        axis.text.y = element_blank())

# Alternative graph.
lamniformes_data |>
  dplyr::select(species, log_cd,log_tl) |>
  ggplot(aes(x = log_cd, y = log_tl)) +
  geom_point(size = 1, shape = 21, fill = "white", colour = "#2A2B3AFF") +
  geom_smooth(method = "lm",col = "#08306b") +
  ggpmisc::stat_poly_eq(use_label(c("eq", "adj.R2", "f", "p")),
                        size = 1.5,label.x = "left",label.y = "top",
                        color = "black") +
  labs(x = "Centrum diameter (log10 cm)", y = "Total length (log10 cm)") +
  facet_wrap(.~species, scales = "free",ncol = 4,nrow = 4) +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "none",
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        strip.text = element_text(face = "bold.italic"),
        aspect.ratio = 1)
  
dist_species <- 
  lamniformes_data |> 
  select(log_tl,log_cd, species) |> 
  pivot_longer(cols = c("log_tl","log_cd"),values_to = "value",names_to = "var") |> 
  ggplot(aes(x = value, y = species)) +
  stat_slab(aes(y = fct_rev(species), fill = species),height = 2, color = "white") +
  scale_fill_manual(values =  col_vec) +
  scale_color_manual(values = col_vec) +
  labs(x = "Centrum diameter (log10 cm)", 
       y = NULL) +
  theme_bw() +
  theme(panel.grid.major.y = element_blank(),
        axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.ticks.y = element_blank(),
        axis.text.y = element_text(face = "italic"),
        legend.position = "none",
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1) +
  facet_wrap(.~var,scales = "free_x")
```

#### Heterogeneous slopes test 

```{r message=FALSE, warning=FALSE}
# Homogeneity of slopes test (additive model).
common_slope <- lm.rrpp(log_tl ~ log_cd + species, SS.type = "III",RRPP = T,iter = 999,data = lamniformes_data)
# Heterogeneity of slopes test (interaction model)
unqiue_slope <- lm.rrpp(log_tl ~ log_cd * species, SS.type = "III",RRPP = T,iter = 999,data = lamniformes_data)

# summary(common_slope)$table # null
# summary(unqiue_slope)$table # alternative

# Which model is better?
# anova(common_slope,unqiue_slope,print.progress = F)$table
model.comparison(common_slope,unqiue_slope,type = "logLik")$table |> 
  kable() |> 
  kable_styling()

# To work with emmeans.
unqiue_slope_em <- lm(log_tl ~ log_cd * species, data = lamniformes_data) # heterogeneity of slopes test (iteraction model)

# Intercepts (predicted values of a when b = 0)
ggarrange(
  emmeans::emmeans(object = unqiue_slope_em, ~ species, at = list(log_cd = 0)) |> 
  as.data.frame() |> 
  suppressMessages() |> 
  ggplot(mapping = aes(x = species,y = emmean, fill = species, colour = species)) +
  geom_point(fill = "white", shape = 21) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),color = col_vec,
                position = position_dodge(0.4), width = .3) +
  scale_fill_manual(values = col_vec) +
  scale_color_manual(values = col_vec) +
  labs(x = NULL, y = "β0 (Intercept)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,hjust = 1,vjust = 1),
        axis.title = element_text(face = "bold"),
        aspect.ratio = 1),
  # Slopes.
  emmeans::emtrends(unqiue_slope_em, specs = "species", var = "log_cd") |> 
  as.data.frame() |> 
  ggplot(mapping = aes(x = species,y = log_cd.trend, fill = species, colour = species)) +
  geom_point(fill = "white", shape = 21) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),color = col_vec,
                position = position_dodge(0.4), width = .3) +
  scale_fill_manual(values = col_vec) +
  scale_color_manual(values = col_vec) +
  labs(x = NULL, y = "β1 (Slope)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,hjust = 1,vjust = 1),
        axis.title = element_text(face = "bold"),
        aspect.ratio = 1),
  # Common legend.
  common.legend = T,legend = "none",label.x = NULL
)

# Compare slopes.
PW <- pairwise(unqiue_slope, groups = lamniformes_data$species, covariate = NULL)

# Select the ones that are significant.
summary(PW, test.type = "dist", confidence = 0.95, stat.table = TRUE)$summary.table  |> 
  as.data.frame() |> 
  # mutate("`Pr > d`" = p.adjust(p = `Pr > d`,method = "fdr"))
  # filter(`Pr > d` <= 0.05) |> 
  mutate(across(1:4, ~ round(.x, digits = 3))) |>
  # filter(Z >= 0)
  # Look at the distribution of effect sizes.
  ggplot(aes(x = Z)) +
  geom_histogram(col = "#25282AFF", fill = "white") +
  geom_vline(xintercept = 0, linetype = 2, col = "#E16A86") +
  labs(x = "Effect size (Z)", y = "Count") +
  theme_bw()
  # knitr::kable(caption = "Pairwise statistics from an analysis of covariance. The dist-test compares LS means 
  #                       (estimated response at mean log-CD) among species.") |> 
  #kable_styling()
```

<br/>

#### Interspecific models

`ML1` -- Interspecific model that includes all sampled lamniform species.
`ML2` -- Interspecific model with reduced taxon coverage; *Alopias* species excluded.
`ML3` -- Interspecific model with reduced taxon coverage; *Alopias* and *Cetorhinus* species excluded.
`ML4` -- Interspecific model with **Lamnidae** species.

```{r}
# Compute trait averages per species.
grp_average <- 
  lamniformes_data |>
  group_by(species) |>
  summarise(log_tl = mean(log_tl),log_cd = mean(log_cd)) |> 
  as.data.frame()

grp_max <- 
  lamniformes_data |>
  group_by(species) |>
  summarise(log_tl = max(log_tl),log_cd = max(log_cd)) |> 
  as.data.frame()

rownames(grp_average) <- grp_average$species

# Lamniformes (all species).
interSM1 <- lm(log_tl ~ log_cd, data = grp_average)
# Lamniformes (excluding Alopidae)
interSM2 <- lm(log_tl ~ log_cd, data = grp_average[c(4:10),])
# Lamniformes (excluding Alopidae and Cetorhinidae)
interSM3 <- lm(log_tl ~ log_cd, data = grp_average[c(4:5,7:10),])
# Lamniformes (excluding all but Lamnidae)
interSM4 <- lm(log_tl ~ log_cd, data = grp_average[c(5,7:9),])

# Compare models.

# ppe(true = 10^grp_average$log_tl,pred = 10^fitted(interSM1))$mean
# ppe(true = 10^grp_average[c(4:10),]$log_tl,pred = 10^fitted(interSM2))$mean
# ppe(true = 10^grp_average[c(4:5,7:10),]$log_tl,pred = 10^fitted(interSM3))$mean
# ppe(true = 10^grp_average[c(5,7:9),]$log_tl,pred = 10^fitted(interSM4))$mean

# How does TL change for one unit increase in CD?

# tidy(interSM1,conf.int = T)
# tidy(interSM2,conf.int = T)
# tidy(interSM3,conf.int = T)
# tidy(interSM4,conf.int = T)
```

```{r fig.cap= "Log-space of lamniform centrum size vs the fossil specimens. This includes all 10 species for which data was available."}
# Plot the log-space of all lamniforms vs the fossil specimens.
bind_rows(
  lamniformes_data |> select(log_cd) |> mutate(grp = "extant"),aptian_shark |> mutate(grp = "extinct")) |> 
  ggplot(mapping = aes(x = log_cd, fill = grp)) +
  geom_histogram(bins = 30, color = "black") +
  scale_fill_manual(values = c("white","#E16A86")) +
  # White shark.
  geom_vline(xintercept = c(0.2379436,0.9315206),linetype = 2) +
  # Basking shark.
  geom_vline(xintercept = c(0.564525,1.181967),linetype = 2, col = "blue") +
  labs(title = NULL,subtitle = NULL,
       y = "Count",
       x = "Centrum diameter (log10 cm)", fill = "Group") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1,legend.position = "none")

bind_rows(
  tidy(interSM1,conf.int = T)[c(1:2,6:7)] |> mutate(model = "ML1"),
  tidy(interSM2,conf.int = T)[c(1:2,6:7)] |> mutate(model = "ML2"),
  tidy(interSM3,conf.int = T)[c(1:2,6:7)] |> mutate(model = "ML3"),
  tidy(interSM4,conf.int = T)[c(1:2,6:7)] |> mutate(model = "ML4")
) |> 
  filter(term == "(Intercept)") |> 
  as.data.frame() |> 
  ggplot(mapping = aes(x = model, y = estimate)) +
  geom_errorbar(aes(x = model, y = estimate,ymin = conf.low, ymax = conf.high),
                width = 0.1, linewidth = 1, col = c("#3B7C70FF","#D01556FF","#CE9642FF","#25282AFF")) +
  geom_point(shape = 21, fill = "white",size = .5) +
  labs(y = "β1 (slope estimate)") +
  # geom_hline(yintercept = 1, lty = 2) +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "none",
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)

# Version 1
lamniformes_data |> 
  mutate(class = case_when(genus == "Alopias" ~ "Alopias",T ~ ""),
         class = case_when(genus != "Alopias" ~ "Non-Alopias", T ~ genus)) |> 
  ggplot(aes(x = log_cd, y = log_tl)) +
  labs(title = NULL,subtitle = NULL,
       y = "Total length (log10 cm)",
       x = "Centrum diameter (log10 cm)", colour = "Species") +
  annotation_logticks() +
  geom_point(aes(colour = class),fill = "white",size = 2, shape = 21) +
  scale_colour_manual(values = c("#2E2D29","#620059")) +
  geom_abline(intercept = coef(interSM1)[1],slope = coef(interSM1)[2], col = "#3B7C70FF") +
  geom_abline(intercept = coef(interSM2)[1],slope = coef(interSM2)[2], col = "#D01556FF") +
  geom_abline(intercept = coef(interSM3)[1],slope = coef(interSM3)[2], col = "#CE9642FF") +
  geom_abline(intercept = coef(interSM4)[1],slope = coef(interSM4)[2], col = "#25282AFF") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "none",
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)

# Version 2.

# data.frame(cd = grp_average_clean$CD,
#            tl = grp_average_clean$TL,
#            labels = grp_average_clean$species) |> 
#   ggplot(aes(x = cd,y = tl)) +
#   geom_point(fill = "black",size = 2, shape = 21) +
#   geom_text_repel(aes(label = labels),
#                   size = 4,
#                   box.padding = unit(0.35, "lines"),
#                   point.padding = unit(0.3, "lines")) +
#   labs(y = "Average total length (log10 cm)",
#        x = "Average centrum diameter (log10 cm)") +
#   annotation_logticks() +
#   geom_abline(intercept = coef(interSM1)[1], slope = coef(interSM1)[2], color = "#3B7C70FF") +
#   geom_abline(intercept = coef(interSM2)[1], slope = coef(interSM2)[2], color = "#D01556FF") +
#   geom_abline(intercept = coef(interSM3)[1], slope = coef(interSM3)[2], color = "#CE9642FF") +
#   geom_abline(intercept = coef(interSM4)[1], slope = coef(interSM4)[2], color = "#25282AFF") +
#   theme_bw() +
#   theme(axis.title = element_text(face = "bold",size = 10),
#         axis.text = element_text(size = 8),
#         axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
#         legend.position = "none",
#         legend.title = element_text(face = "bold"),
#         legend.text = element_text(face = "italic"),
#         plot.title = element_text(face = "bold",size = 12),
#         plot.subtitle = element_text(size = 10),
#         aspect.ratio = 1)
```

### Fossil Predictions (II)

Total length predictions are here computed based on our intraspecific regression equations.

```{r}
# LM fits.
species_fit <- 
  lamniformes_data |> 
  group_by(species) |> 
  do(model = fit_regression(.))

# Provide taxon names to each sub-list.
for (i in seq_along(species_fit$model)) {
  names(species_fit$model)[[i]] <- sort(species_fit$species)[i]
}

# Compute evaluation metrics.
summary_stats <- 
  matrix(nrow = 10,ncol = 2, dimnames = list(species_fit$species,c("MSE","R.squared")))

for (i in 1:10) {
  summary_stats[i,1] <- round(mean(summary(species_fit$model[[i]])$residuals^2),digits = 3)
  summary_stats[i,2] <- round(summary(species_fit$model[[i]])$adj.r.squared,digits = 3)
}

# Predictions for the Darwin Formation vertebral centra based on extant species-level equations.
species_prediction <- vector("list",length = 10)

names(species_prediction) <- 
  species_fit$species

for(i in seq_along(species_fit$model)) {
  preds <- predict(species_fit$model[[i]],newdata = aptian_shark, type = "response",interval = "prediction",level = .95)
  species_prediction[[i]] <- 10^preds # anti-log
}

# Add specimen label.
species_prediction <- purrr::map(species_prediction, ~as.data.frame(.x),.progress = T)
species_prediction <- purrr::map(species_prediction, ~mutate(.x, specimens = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29")))
```

#### Model performance and auxiliary results

The performance of each species-level model is evaluated using the **Mean Absolute Error (MAE)** and **Percent Prediction Error (PPE)**. This MAE is a measure of **goodness-of-fit** and represents the average absolute difference between the predicted (fitted) values and the actual target values (in this context TL). PPE is

**MAE** is given by:

$$MAE = \sum_{i=1}^n |y_i - x_i| \over n$$

```{r}
# Model performance by species.
prepare_split <- 
  lamniformes_data |> group_by(species)

prepare_split_frame <- 
  group_split(prepare_split) |> lapply(as.data.frame)

QR <- vector("list", length = 10)

for(i in seq_along(species_fit$model)) {
  preds <- predict(species_fit$model[[i]],newdata = data.frame(log_cd = prepare_split_frame[[i]]$log_cd), type = "response",interval = "prediction",
                   level = .95)
  QR[[i]] <- preds
}

names(QR) <- species_fit$species

# Add TL values for each species back to the predicted TL and CI dataframes. 
predicted_actual_set <- lapply(seq_along(QR), function(x) {
  lsName <- names(QR)[x]
  log_tl <- lamniformes_data$log_tl[lamniformes_data$species == lsName]
  cbind(QR[[x]], log_tl = log_tl)
})

names(predicted_actual_set) <- species_fit$species

maeR <- predicted_actual_set |> lapply(as.data.frame)

# MAE.
lapply(maeR, function(x) { Metrics::mae(x$log_tl, x$fit) }) |> 
  unlist() |> 
  as.data.frame() |> 
  setNames(nm = "mae") |> 
  rownames_to_column(var = "species") |> 
  mutate("mae" = round(mae,digits = 2)) |> 
  kable() |> 
  kable_styling()

# PPE
lapply(maeR, function(x) { MASSTIMATE::ppe(10^x$log_tl, 10^x$fit)$mean }) |> 
  unlist() |> 
  as.data.frame() |> 
  setNames(nm = "PPE") |> 
  rownames_to_column(var = "species") |> 
  # mutate("|PPE%|" = round(PPE,digits = 2)) |> 
  kable() |> 
  kable_styling()
```

<br/>

#### Full GWS {#sec-gws}

<!-- Analysis using all GWS records -->

```{r fig.cap="Relationship between centrum diameter and total length in C. carcharias. Data is based on records from Wintner & Cliff (1999) and Cailliet et al. 1985. The estimated TL along with 95% prediction intervals of the cardabiodontid specimens are added to the graph", warning=FALSE, message=FALSE}

# Prediction intervals.
gws_pi <- 
  predict(object = slope_intercept$model[[5]],newdata = data.frame(log_cd = carcharodon_carcharias_complete$log_cd),
          interval = "prediction") |> as.data.frame()
# Plot.
p2 <- 
  carcharodon_carcharias_complete |> 
  ggplot(mapping = aes(x = log_cd, y = log_tl)) +
  geom_point(aes(colour = reference), shape = 19) +
  annotation_logticks() +
  scale_x_continuous(limits = c(min(carcharodon_carcharias_complete$log_cd),max(aptian_shark$log_cd)+0.02)) +
  scale_y_continuous(limits = c(min(carcharodon_carcharias_complete$log_tl),log10(max(species_prediction$`Carcharodon carcharias`$upr)))) +
  geom_smooth(method = "lm",se = F, colour = "black") +
  ggpmisc::stat_poly_eq(use_label(c("eq", "adj.R2", "f", "p", "n")),
                        size = 4.0,label.x = "left",label.y = "top",
                        color = "black") +
  scale_fill_manual(values =  c("#E16A86","#08306b")) +
  scale_color_manual(values = c("#E16A86","#08306b")) +
  # Prediction intervals.
  geom_line(aes(y = gws_pi[,"lwr"]), col = "black", linetype = "dashed") +
  geom_line(aes(y = gws_pi[,"upr"]), col = "black", linetype = "dashed") +
  # Add fossil predictions to base plot.
  geom_point(data = aptian_shark,mapping = aes(x = log_cd, y = log10(species_prediction$`Carcharodon carcharias`$fit)),
             colour = "black", fill = "white",pch = 22, size = 2) +
  # Labels.
  labs(x = "Centrum diameter (log10 cm)", y = "Total length (log10 cm)") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_blank(),
        aspect.ratio = 1)

p3 <- p2 + 
  geom_errorbar(data = species_prediction$`Carcharodon carcharias`, 
                aes(x = aptian_shark$log_cd,y = log10(fit), ymin = log10(lwr),ymax = log10(upr))
                ,color = "black")

print(p3)

# PPE.
complete_gws_ppe <-
  ppe(true = 10^carcharodon_carcharias_complete$log_tl,
      pred = 10^as.vector(fitted(slope_intercept$model[[5]])))
```

<!--Comparative species predictions-->

```{r}
data.frame(Reduce(rbind, species_prediction)) |>
  as_tibble() |> 
  mutate(taxon = rep(species_fit$species,each = 5)) |>
  mutate(specimens = fct_relevel(specimens,c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29"))) |> 
  as.data.frame() |> 
  knitr::kable(caption = "Specimen-level predictions based on extant intraspecific regression equations (see Figure 15).") |> 
  kableExtra::kable_styling()
```

```{r fig.cap = "Total length estimates of the Darwin Formation cardabiodontid shark. Individual panel corresponds to a unique specimen and show the estimated (back-transformed) values along with 95% prediction intervals generated from our lamniform species-level data subsets."}
# Based on all lamniform species.
data.frame(Reduce(rbind, species_prediction)) |>
  as_tibble() |> 
  mutate(taxon = rep(species_fit$species,each = 5)) |>
  mutate(specimens = fct_relevel(specimens,c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29"))) |> 
  # Plot.
  ggplot(aes(y = taxon, x = fit, colour = taxon)) +
  geom_errorbar(aes(xmin = lwr, xmax = upr),
                width = 0.3, linewidth = 1.3,
                position = position_dodge(width = 0.9)) +
  geom_point(size = 2, fill = "white", shape = 21,
             position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = col_vec) +
  scale_color_manual(values = col_vec) +
  coord_flip() +
  facet_wrap(.~specimens,ncol = 5) +
  labs(x = "Total length (cm)",y = "",
       title = NULL,
       subtitle = NULL,
       color = "Models") +
  theme_bw() +
  theme(title = element_text(face = "bold"),
        axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic",size = 8),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        strip.text = element_text(face = "bold"),
        aspect.ratio = 1.8)
```

#### Lamnidae overview

```{r message=FALSE,warning=FALSE}
# Plot the log-space of C. carcharias centrum size vs the fossil specimens.
bind_rows(lamnidae_data |> 
  filter(species == "Carcharodon carcharias") |> 
  select(log_cd,species,reference),aptian_shark |> mutate(species = "NTM P",reference = "NTM P")) |> 
  ggplot(mapping = aes(x = log_cd, fill = reference)) +
  geom_histogram(bins = 30) +
  scale_fill_manual(values = c("#B83A4B","#2E2D29","#017E7C")) +
  theme_bw() +
  theme(axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold",angle = 90,hjust = 0.5,vjust = .5),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "plain"),
        aspect.ratio = 1) +
  labs(title = "Log-10 space",
       subtitle = NULL,
       x = "Centrum diameter (log10 cm)",
       y = "Count", fill = "")

# Check for negative values.

# lamnidae_data |> 
#   select(species,contains("log_")) |>
#   filter(log_cd < 0.0)

# Lamnidae species CD-TL relationship.
lamnidae_data |> 
  ggplot(aes(x = log_cd, y = log_tl)) +
  geom_point(aes(colour =  species), fill = "white",size = 2, shape = 21) +
  geom_smooth(aes(colour = species),method = "lm",show.legend = F,se = F) +
  annotation_logticks() +
  # Species colors.
  scale_color_manual(values = c("#E76F51","#1D3557", "#F0A8C0FF", "#6D6875")) +
  scale_fill_manual(values =  c("#E76F51","#1D3557", "#F0A8C0FF", "#6D6875")) +
  labs(x = "Centrum diameter (log10 cm)",
       y = "Total length (log10 cm)",
       colour = "Taxa",
       title = "Lamnidae",
       subtitle = "Centrum diameter versus total length") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)
```

#### Residual analysis

```{r}
# Linear fit.
lamnidae_model <- lm(log_tl ~ log_cd, lamnidae_data)

# Average Z-score for predicted TL.
zScores = mean((log10(pred_tl$fit) - mean(predict(lamnidae_model)))/sd(predict(lamnidae_model)))

# Plot results.
lamnidae_data |> 
  ggplot(mapping = aes(x = predict(lamnidae_model))) +
  geom_histogram(bins = 30, color = "black", fill = "white") +
  #xlim(c(1.949668,4.080516)) +
  geom_vline(xintercept = mean(predict(lamnidae_model)), linewidth = 1, linetype="dashed",colour = "#E16A86") +
  geom_vline(xintercept = zScores, linewidth = 1, linetype="dashed", colour = "#08306b") +
  # Two standard deviations around the mean.
  # geom_vline(xintercept = c((lamnidae_mean - 2 * lamnidae_sd),(lamnidae_mean + 2 * lamnidae_sd)),colour = "black") +
  labs(title = NULL,subtitle = NULL,
       y = "Count",
       x = "Predicted total length (log10)") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1,legend.position = "none")
```

#### Alopidae overview

```{r warning=FALSE,message=FALSE}
# Alopias log-space and Aptian specimens.
bind_rows(alopiidae_data |> mutate(status = "Extant"), 
          aptian_shark |> mutate(status = "Fossil")
) |> 
  ggplot(aes(x = log_cd, fill = status)) +
  geom_histogram() +
  scale_fill_manual(values = c("#E16A86","black")) +
  theme(aspect.ratio = 1/3,
        plot.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold")) +
  labs(title = "Log-10 space", fill = NULL,
       subtitle = "Alopias-species centrum diameter versus NTM P",
       x = "Centrum diameter (log10, cm)", y = "Count")

# Only Alopias species.
alopiidae_data |> 
  ggplot(aes(x = log_cd, y = log_tl)) +
  geom_point(aes(color = species), fill = "white",size = 2, shape = 21) +
  # Averages.
  geom_point(data = grp_average[c(1:3),], aes(x = log_cd, y = log_tl, colour = species), shape = 19,size = 4,show.legend = F) +
  annotation_logticks() +
  geom_smooth(aes(colour = species),method = "lm",show.legend = F,se = F) +
  scale_color_manual(values = c("#264653", "#2A9D8F", "#E9C46A")) +
  scale_fill_manual(values =  c("#264653", "#2A9D8F", "#E9C46A")) +
  labs(title = "Alopiidae — Thresher sharks",subtitle = "Centrum diameter versus total length",
       y = "Total length (log10 cm)",
       x = "Centrum diameter (log10 cm)", colour = "Species") + 
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)
```

### Fossil Predictions (III)

Here we compute **TL** estimates based on our interspecific regression equations.

```{r}
# Predictions on interspecific models.
IS.I <- 
  cbind(
  # Model 1.
  10^predict(object = interSM1,newdata = aptian_shark,interval = "prediction"),
  10^predict(object = interSM1,newdata = aptian_shark,interval = "confidence")[,2:3]) |> 
  as.data.frame() |> 
  unname() |> 
  setNames(nm = c("fit",".pred_lower",".pred_upper",".conf_lower",".conf_upper")) |> 
  mutate(specimen = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29")) |> 
  mutate(model = "Model I")

IS.II <- 
  cbind(
  # Model 2.
  10^predict(object = interSM2,newdata = aptian_shark,interval = "prediction"),
  10^predict(object = interSM2,newdata = aptian_shark,interval = "confidence")[,2:3]) |> 
  as.data.frame() |> 
  unname() |> 
  setNames(nm = c("fit",".pred_lower",".pred_upper",".conf_lower",".conf_upper")) |> 
  mutate(specimen = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29")) |> 
  mutate(model = "Model II")

IS.III <-
  cbind(
  # Model 3.
  10^predict(object = interSM3,newdata = aptian_shark,interval = "prediction"),
  10^predict(object = interSM3,newdata = aptian_shark,interval = "confidence")[,2:3]) |> 
  as.data.frame() |> 
  unname() |> 
  setNames(nm = c("fit",".pred_lower",".pred_upper",".conf_lower",".conf_upper")) |> 
  mutate(specimen = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29")) |> 
  mutate(model = "Model III")

IS.IIII <-
  cbind(
  # Model 4.
  10^predict(object = interSM4,newdata = aptian_shark,interval = "prediction"),
  10^predict(object = interSM4,newdata = aptian_shark,interval = "confidence")[,2:3]) |> 
  as.data.frame() |> 
  unname() |> 
  setNames(nm = c("fit",".pred_lower",".pred_upper",".conf_lower",".conf_upper")) |> 
  mutate(specimen = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29")) |> 
  mutate(model = "Model IV")

# Plot results.
bind_rows(IS.I,IS.II,IS.III,IS.IIII) |> 
  mutate(specimen = fct_relevel(specimen,c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29"))) |> 
  # Plot.
  ggplot(aes(y = model, x = fit, colour = model)) +
  geom_errorbar(aes(xmin = .pred_lower, xmax = .pred_upper),
                width = 0.3, linewidth = 1.3,
                position = position_dodge(width = 0.9)) +
  geom_point(size = 2, fill = "white", shape = 21,
             position = position_dodge(width = 0.9)) +
  coord_flip() +
  
  scale_fill_manual(values =  c("#3B7C70FF","#D01556FF","#CE9642FF","#25282AFF")) +
  scale_color_manual(values = c("#3B7C70FF","#D01556FF","#CE9642FF","#25282AFF")) +
  facet_wrap(.~specimen,ncol = 5) +
  labs(x = "Total length (cm)",y = "",
       title = NULL,
       subtitle = NULL,
       color = "Interspecific models") +
  theme_bw() +
  theme(title = element_text(face = "bold"),
        axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        legend.text = element_text(size = 7),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        strip.text = element_text(face = "bold"),
        aspect.ratio = 1.8)

# PPE.

# intervals(xvar = IS.I$fit)[c(1,7:8)]
# intervals(xvar = IS.II$fit)[c(1,7:8)]
# intervals(xvar = IS.III$fit)[c(1,7:8)]
# intervals(xvar = IS.IIII$fit)[c(1,7:8)]
```

## Mesozoic data

```{r message=FALSE, warning=FALSE}
# Minor feature engineering.
mesozoic_data_subset <- 
  mesozoic_data |> 
  mutate(log_cd = log10(centrum_diameter),
         log_tl = log10(total_length)) |> 
  filter(species != "Cardabiodon ricki")

# Visualize.
meso_01 <-
  mesozoic_data_subset |> 
  ggplot(aes(x = log_cd, y = log_tl, color = family)) +
  geom_point() +
  geom_smooth(aes(fill = family),method = "lm", se = F, color = "#08306b") +
  scale_fill_manual(values =  c("#D01556FF","#175E54")) +
  scale_color_manual(values = c("#D01556FF","#175E54")) +
  stat_poly_eq(use_label(c("eq", "adj.R2", "f", "p", "n")),
               size = 4.0,label.x = "left",label.y = "top") +
  # geom_smooth(method = "rlm", se = FALSE, color = "#027EB6",linetype = 2) + 
  labs(x = "Centrum diameter (log10 cm)",
       y = "Total length (log10 cm)") +
  geom_label_repel(aes(label = label),size = 2,
                  box.padding = unit(0.35, "lines"),
                  point.padding = unit(0.5, "lines"),
                  segment.color = 'grey50') + 
  # facet_wrap(.~family, scale = "free") +
  theme_bw() +
  theme(title = element_text(face = "bold"),
        axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.x = element_text(hjust = 1,vjust = 1),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_text(face = "bold"),
        legend.position = "none",
        aspect.ratio = 1)

# Fit linear model.
stem_fit_01 <- lm(log_tl ~ log_cd, data = mesozoic_data_subset)
stem_fit_02 <- lm(log_tl ~ log_cd, data = mesozoic_data_subset,subset = mesozoic_data_subset$family == "non-Anacoracidae")
# stem_fit |> summary()

# OLS prediction.
pred_stem_ols_01 <- 
  10^predict(stem_fit_01, newdata = aptian_shark,interval = "prediction",level = 0.95) |> 
  bind_cols(specimen = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29")) |> 
  as.data.frame() |> 
  add_column(status = rep("OLS: complete data",5))

pred_stem_ols_02 <- 
  10^predict(stem_fit_02, newdata = aptian_shark,interval = "prediction",level = 0.95) |> 
  bind_cols(specimen = c("NTM P 922-33","NTM P 922-32","NTM P 9382", "NTM P 922-37","NTM P 922-29")) |> 
  as.data.frame() |> 
  add_column(status = rep("OLS: non-Anacoracidae",5))

# Compare OLS vs RLM prediction results.
stem_all_predictions <- 
  bind_rows(pred_stem_ols_01,pred_stem_ols_02) |> 
  reshape::melt(c("fit","lwr","upr","specimen","status"))

# Plot.
meso_02 <- stem_all_predictions |> 
  ggplot(aes(y = specimen, x = fit, colour = status)) +
  geom_errorbar(aes(xmin = lwr, xmax = upr),
                width = 0.3, linewidth = 1.3,
                position = position_dodge(width = 0.9)) +
  geom_point(size = 3, fill = "white", shape = 21,
             position = position_dodge(width = 0.9)) +
  scale_fill_manual(values =  c("#08306b","#D01556FF")) +
  scale_color_manual(values = c("#08306b","#D01556FF")) +
  geom_vline(xintercept = c(514.3853,695.5629), col = c("#08306b","#D01556FF"), linetype = c(2,2)) + 
  theme_bw() +
  labs(x = "Total length (cm)",y = "", color = "Model") +
  theme(legend.position = "none",
        title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "plain"),
        strip.text = element_blank(),
        axis.text.y = element_blank(),
        axis.text = element_text(size = 8),
        axis.title.x = element_text(vjust = -3),
        plot.caption = element_text(hjust = 0),
        aspect.ratio = 1)
  # facet_wrap(~factor(specimen,c("NTM P922-32","NTM P922-29",
  #                               "NTM P 922-37","NTM P 922-33")),scales = "free_y",ncol = 2)

ggarrange(meso_01,meso_02,ncol = 2,align = "hv")
```

## Phylogenetic least squares

<!--Tree from Stein et al. 2018. Downloaded from: https://vertlife.org/sharktree/-->

```{r}
# Genetic Data Only Tree.
shark_tree <- readNexus(file = "phylogeny/500treePLtrees.nex")
# MCC Tree.
mcc_tree <- maxCladeCred(x = shark_tree,rooted = T) # Genetic Data Only Trees.

grp_average_clean <- data.frame(CD = setNames(grp_average$log_cd,gsub(" ", "_", grp_average$species)),
                                TL = setNames(grp_average$log_tl,gsub(" ", "_", grp_average$species)),
                                species = gsub(" ", "_", grp_average$species))

# layout(matrix(c(1:2),nrow = 2),respect = T)

# Drop tips.
lamn_tree <- treedata(phy = mcc_tree,data = grp_average_clean, warnings = FALSE)$phy
plotTree(ladderize(lamn_tree))

# Alternative lamniform MCC topology.
modified_tree <- readNexus(file = "phylogeny/mccTree_alopias.nex") # Make Alopias monophyletic.
# Drop-tips.
modified_lamn <- treedata(phy = modified_tree,data = grp_average_clean, warnings = FALSE)$phy

# Calculates distances to the root
# adephylo::distRoot(modified_lamn) 

# All tips should be 115.9955 Ma from the root to be ultrametric
# A. superciliosus needs to reduce by 132.6635 - 115.9955 = 16.668
modified_lamn$edge.length[6] <- 88.436256 - 16.668

plotTree(ladderize(modified_lamn)) # re-check tree
distRoot(modified_lamn) # re-calculates distances to the root, should all be equal

# PGLS (via RRPP).
grp_pgls <- rrpp.data.frame(CD = setNames(grp_average_clean$CD,gsub(" ", "_", grp_average_clean$species)),
                            TL = setNames(grp_average_clean$TL,gsub(" ", "_", grp_average_clean$species)),
                            species = gsub(" ", "_", grp_average_clean$species),
                            ecology = c(rep("macrophagous",5),rep("filter-feeding",1),rep("macrophagous",4)))

grp_pgls$CovA <- vcv.phylo(lamn_tree)
grp_pgls$CovB <- vcv.phylo(modified_lamn) # Modified.

# Fit model for the first topology.
pgls_shark <- lm.rrpp(f1 = TL ~ CD,data = grp_pgls, Cov = grp_pgls$CovA,  print.progress = FALSE)
anova(pgls_shark)

nonPhylo_shark <- lm.rrpp(f1 = TL ~ CD,data = grp_pgls,  print.progress = FALSE)

model.comparison(pgls_shark,nonPhylo_shark,type = "logLik") # PGLS is better.

# Fit model for the second topology.
pgls_shark_02 <- lm.rrpp(f1 = TL ~ CD,data = grp_pgls, Cov = grp_pgls$CovB,  print.progress = FALSE)

# Generalized Least-Squares (via OLS projection)
# tree_ggplot <- 
#   ggtree(lamn_tree) + 
#   geom_tiplab(as_ylab=TRUE, color='black') +
#   theme(aspect.ratio = 1)

pgls_plot <- 
  data.frame(cd = grp_average_clean$CD,
           tl = grp_average_clean$TL,
           labels = grp_average_clean$species) |> 
  ggplot(aes(x = cd,y = tl)) +
  geom_point(fill = "black",size = 2, shape = 21) +
  geom_text_repel(aes(label = labels),
                  size = 4,
                  box.padding = unit(0.35, "lines"),
                  point.padding = unit(0.3, "lines")) +
  labs(y = "Average total length (log10)",
       x = "Average centrum diameter (log10)") +
  annotation_logticks() +
  geom_abline(intercept = coef(pgls_shark_02)[1], slope = coef(pgls_shark_02)[2], color = "#175E54", linetype = 2) +
  geom_abline(intercept = coef(pgls_shark)[1], slope = coef(pgls_shark)[2], color = "#B2182B") +
  geom_abline(intercept = coef(nonPhylo_shark)[1], slope = coef(nonPhylo_shark)[2], color = "blue") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.position = "none",
        legend.title = element_text(face = "bold"),
        legend.text = element_text(face = "italic"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)
```

### Phylogenetic signal

```{r}
# Signal including Blombergs K and Effect Size.

# zCD <- geomorph::physignal.z(A = setNames(grp_average_clean[,"CD"],gsub(" ","_",grp_average_clean$species)),
#                              phy = lamn_tree,lambda = "front", iter=999)[1]
# 
# zTL <- geomorph::physignal.z(A = setNames(grp_average_clean[,"TL"],gsub(" ","_",grp_average_clean$species)),
#                              phy = lamn_tree,lambda = "front", iter=999)[1]
# 
# set.seed(1)

# cbind(
  # Blomberg K.
#   phytools::phylosig(tree = lamn_tree,
#                      x = setNames(grp_average_clean[,"CD"],gsub(" ","_",grp_average_clean$species)),
#                      test = T)[1:2] |> unlist(),
#   phytools::phylosig(tree = lamn_tree,
#                      x = setNames(grp_average_clean[,"TL"],gsub(" ","_",grp_average_clean$species)),
#                      test = T)[1:2] |> unlist()
# ) |> 
#   magrittr::set_colnames(c("CD", "TL")) |>
#   as.data.frame() |> 
#   add_row(CD = unlist(unname(zCD)), TL = unlist(unname(zTL))) |> 
#   t() |> 
#   magrittr::set_colnames(c("K", "P","Z")) |> 
#   knitr::kable(caption = "Phylogenetic signal in CD and TL") |>
#   kableExtra::kable_styling()

# Sensitivity test (exclude Alopias species)
# exclude_alopias_d <- grp_average_clean[c(4:10),]
# exclude_alopias_t <- treedata(phy = mcc_sgTree,data = exclude_alopias_d, warnings = FALSE)$phy

# Effect size.
# zCD_alopias <- geomorph::physignal.z(A = setNames(exclude_alopias_d[,"CD"],gsub(" ","_",exclude_alopias_d$species)),
#                              phy = exclude_alopias_t,lambda = "front", iter=999)[1]
# 
# zTL_alopias <- geomorph::physignal.z(A = setNames(exclude_alopias_d[,"TL"],gsub(" ","_",exclude_alopias_d$species)),
#                              phy = exclude_alopias_t,lambda = "front", iter=999)[1]
# 
# set.seed(2)

# Blomberg K.
# cbind(
#   phytools::phylosig(tree = exclude_alopias_t,
#                      x = setNames(exclude_alopias_d[,"CD"],gsub(" ","_",exclude_alopias_d$species)),
#                      test = T)[1:2] |> unlist(),
#   phytools::phylosig(tree = exclude_alopias_t,
#                      x = setNames(exclude_alopias_d[,"TL"],gsub(" ","_",exclude_alopias_d$species)),
#                      test = T)[1:2] |> unlist()
# )
```

## Modern body-size ranges

```{r message=FALSE, warning=FALSE}
# Body size ranges.
size_trait <- 
  openxlsx::read.xlsx(xlsxFile = "data/lamniformes-body-ranges.xlsx") |> 
  mutate(min = as.numeric(min),
         max = as.numeric(max))

rownames(size_trait) <- size_trait$species

bzR <- data.frame(min = setNames(size_trait$min, size_trait$species),
                  max = setNames(size_trait$max, size_trait$species),
                  species = size_trait$species)

# Complete lamniform tree.
# lamniformTree <- geiger::treedata(phy = mcc_sgTree,data = bzR, warnings = FALSE)$phy # 14 lamnifom taxa

xe <- c(setNames(size_trait$min, size_trait$species),
        setNames(size_trait$max, size_trait$species)
        )

xt <- xe[order(names(xe))][-c(25:26)] # Remove Odontaspis_ferox (Missing from tree)

# Plot.
# par(mfrow=c(1,2))
# plotTree(lamniformTree,mar=c(5.1,1.1,2.1,0.1)); axisPhylo(1) 
# par(mar=c(5.1,0.1,2.1,1.1))
# boxplot(xt~factor(names(xt),levels=lamniformTree$tip.label),horizontal=TRUE,xlab = "Body size (cm)",
#     axes=F,xlim=c(1,Ntip(lamniformTree)))
# 
# axis(1)
```

## GWS vertebrae counts

Make sure you note the number of verts was only the total amount we sampled, they had more verts but we stopped sampling around the caudal peduncle. You might want to make sure that is clear.

$$Y_i \sim \text{Poisson}(\lambda_i)$$

```{r warning=FALSE}
# Plot data
gws_long_format <- 
  bind_rows(natanson_data[[3]],natanson_data[[4]],
              natanson_data[[5]],natanson_data[[6]]) |> 
  janitor::clean_names()

column_label <- tibble(
  vertebrae = Inf,
  lateral_diameter_mm = Inf,
  specimen = unique(gws_long_format$specimen),
  label = c("Column specimen 144.9 cm FL, Male",
            "Column specimen 152.0 cm FL, Male",
            "Column specimen 380.9 cm FL, Male",
            "Column specimen 331.0 cm FL, Male"),
  x.pos = c(25,25,30,25),
  y.pos = c(15.0,19.0,50,40))

largest_centra <- gws_long_format |> 
  group_by(specimen) |> 
  slice_max(lateral_diameter_mm,na_rm = T) |> 
  ungroup()

gws_long_format |> 
  ggplot(aes(x = vertebrae, y = lateral_diameter_mm)) + 
  geom_point(shape = 21, fill = "white", color = "black", size = 1.5) +
  labs(y = "Lateral diameter (mm)", x = "Vertebral number",
       title = NULL,
       subtitle = NULL) +
  facet_wrap(.~specimen,scales = "free") +
  geom_text(data = column_label,mapping = aes(x = x.pos, y = y.pos, label = label),size = 3, color = "black") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold",size = 12),
        plot.subtitle = element_text(size = 10),
        aspect.ratio = 1)

# Four white shark individuals with size, vertebrae count, and centrum diameter.
gws_natanson <-
  data.frame(n_ver = lapply(natanson_data, nrow)[3:6] |> unlist(),
             max_d = lapply(natanson_data, function(x) max(x[[2]],na.rm = T))[3:6] |> unlist()/10,
             max_t = lapply(natanson_data, function(x) max(x[[3]],na.rm = T))[3:6] |> unlist()/10,
             fl_cm = lapply(natanson_data, function(df) df[[4]][1])[3:6] |> unlist(),
             sex = c("Male","Male","Male","Male"))

gws_natanson <- gws_natanson |> select(-5)

gws_natanson_log <- 
  cbind(gws_natanson[,1],apply(gws_natanson[,2:4],2,log10)) |>
  as.data.frame() |>
  rename("n_ver" = V1)

# Poisson model.
pos_model <- glm(n_ver ~ fl_cm + max_d + max_t, family = "poisson",data = gws_natanson_log)
summary(pos_model)

tidy(pos_model,conf.int = TRUE) |> 
  knitr::kable(caption = "GLM results") |>
  kable_styling(bootstrap_options = "striped",full_width = F)

nS <- 
  natanson_supplementary |>
  mutate(across(4:6, ~ . /10)) |> # Convert to cm.
  group_by(Specimen) |> 
  summarise(n_ver = max(Column),
            max_d = max(`Dorsal.Diameter.(mm)`),
            max_t = max(`Thickness.(mm)`),
            fl_cm = max(`FL.(cm)`)) |> 
  as.data.frame() |> 
  column_to_rownames(var = "Specimen")

nST <- 
  bind_rows(gws_natanson_log,nS |> mutate(across(2:4, ~ log10(.x)))) |> 
  mutate(Species = c(rep("White shark",4),
                     rep("Common Thresher",2),
                     rep("Shortfin Mako",3),
                     rep("Porbeagle",3))
         )

pos_model_all <- glm(n_ver ~ fl_cm + max_d + max_t, family = "poisson",data = nST)
summary(pos_model_all)

# Correlation test.
cor(gws_natanson$n_ver,gws_natanson$max_d,method = "spearman")
cor(gws_natanson$n_ver,gws_natanson$fl_cm,method = "spearman")
cor(gws_natanson$max_d,gws_natanson$fl_cm,method = "spearman")
cor(gws_natanson$max_t,gws_natanson$fl_cm,method = "spearman")

# All species,
cor(bind_rows(nS,gws_natanson)$n_ver,bind_rows(nS,gws_natanson)$max_d,method = "spearman")
cor(bind_rows(nS,gws_natanson)$n_ver,bind_rows(nS,gws_natanson)$fl_cm,method = "spearman")
cor(bind_rows(nS,gws_natanson)$max_d,bind_rows(nS,gws_natanson)$fl_cm,method = "spearman")
cor(bind_rows(nS,gws_natanson)$max_t,bind_rows(nS,gws_natanson)$fl_cm,method = "spearman")
```

## Re-estimating extinct species

**OMNH 68860: CD = 11 cm)** -- Original estimate by [@Frederickson2015] was 630 and 660 cm TL.

**FHSM VP-2187 (CD = 8.7 cm):** -- Original estimate by [@Shimada1997b] was 500 cm Tl. Refined estimate by Newbrey et al. of 5.3 m.

```{r warning=FALSE, message=FALSE}
# First use the regression equation by Gottfried et al. (1996) to confirm the result in Frederickson et al. 2015.
# The CD here is in mm units and result in m.

# 0.22 + 0.058 * 110

# Use Lamnidae interspecific model to re-estimate the TL.

# 10^predict(object = interSM4,newdata = data.frame(log_cd = log10(11.0)),interval = "prediction")
# 10^predict(object = interSM4,newdata = data.frame(log_cd = log10(8.7)), interval = "prediction")

# Cretaceous lamniform species.
stem_species <- 
  data.frame(log_cd = log10(mesozoic_data$centrum_diameter/10))

10^predict(object = interSM3,newdata = stem_species,interval = "prediction") |> 
  data.frame() |> 
  add_column("specimen" = mesozoic_data$label) |> 
  mutate(CD= mesozoic_data$centrum_diameter/10) |> 
  add_row(fit = 592.5426, lwr = 493.8996, upr = 710.8868,specimen = "NTM P 922-32", CD = 12.6) |> 
  # Order by age
  mutate(specimen = factor(specimen, levels = c("NTM P 922-32","OMNH 68860","KUVP 16343","WAM 96.4.45.108",
                                                "MMSP CPC 3063","MMSP CPC 3068",
                                                "IGVR 91032","MPPSA-IGVR 36371","FHSM VP-2187","MSNPV 20407–20417",
                                                "LACM 128126","USNM 425665","LACM 135929","LACM 143537",
                                                "LACM 120090","SDSM 34975","SDSM 47682","SDSM 47683"))) |> 
  ggplot(aes(x = specimen, y = fit)) +
  geom_errorbar(aes(ymin = lwr, ymax = upr),
                width = 0.1,linewidth = 1.3) +
  geom_point(size = 3, shape = 19,colour = "black") +
  # Original points.
  geom_point(data = mesozoic_data,aes(x = label, y = total_length/10), size = 3, shape = 19, colour = "#B83A4B") +
  # CD label.
  #geom_text(aes(x = specimen, y = fit + 200, label = paste0("CD=",as.character(CD),"cm")),size = 6, col = "black") +
  geom_rect(xmin = 0.7, xmax = 1.3, ymin = 450, ymax = 750,lty = 2,alpha = 0, color = "black") +
  # ‘small,’ 1‒200 cm; ‘medium,’ 201‒400 cm; ‘large,’ 401‒600 cm; and ‘gigantic,’ >600 cm). 
  geom_hline(yintercept = c(201,400,600), lty = c(2,2,1), col = "#08306b") +
  theme_bw() +
  labs(x = NULL, y = "Estimated total length (cm)",title = NULL) +
  theme(title = element_text(face = "bold"),
        axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.x = element_text(hjust = 1,vjust = 1,angle = 45),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold",size = 12),
        aspect.ratio = 1/2)

# Exclude Anacoracids.
10^predict(object = interSM3,newdata = stem_species,interval = "prediction") |> 
  data.frame() |> 
  add_column("specimen" = mesozoic_data$label) |> 
  mutate(CD= mesozoic_data$centrum_diameter/10) |> 
  add_row(fit = 592.5426, lwr = 493.8996, upr = 710.8868,specimen = "NTM P 922-32", CD = 12.6) |>
  as_tibble() |> 
  # Anacoracidae
  filter(!specimen %in% c("USNM 425665","LACM 135929","LACM 143537", "LACM 120090", "SDSM 34975", "SDSM 47682", "SDSM 47683")) |> 
  # Order by Age,
  mutate(specimen = factor(specimen, levels = c("NTM P 922-32",
                                                "OMNH 68860",
                                                "KUVP 16343",
                                                "WAM 96.4.45.108",
                                                "IGVR 91032", 
                                                "MPPSA-IGVR 36371", 
                                                "FHSM VP-2187", 
                                                "MSNPV 20407–20417", 
                                                "LACM 128126"))) |> 
  ggplot(aes(x = specimen, y = fit)) +
  geom_errorbar(aes(ymin = lwr, ymax = upr),
                width = 0.1,linewidth = 1.3) +
  geom_point(size = 3, shape = 19,colour = "black") +
  # Original points.
  geom_point(data = mesozoic_data |> filter(family != "Anacoracidae"),aes(x = label, y = total_length/10), size = 3, shape = 19, colour = "#B83A4B") +
  # CD label.
  #geom_text(aes(x = specimen, y = fit + 200, label = paste0("CD=",as.character(CD),"cm")),size = 6, col = "black") +
  geom_rect(xmin = 0.7, xmax = 1.3, ymin = 450, ymax = 750,lty = 2,alpha = 0, color = "black") +
  # ‘small,’ 1‒200 cm; ‘medium,’ 201‒400 cm; ‘large,’ 401‒600 cm; and ‘gigantic,’ >600 cm). 
  geom_hline(yintercept = c(201,400,600), lty = c(2,2,1), col = "#08306b") +
  theme_bw() +
  labs(x = NULL, y = "Estimated total length (cm)",title = NULL) +
  theme(title = element_text(face = "bold"),
        axis.title = element_text(face = "bold",size = 10),
        axis.text = element_text(size = 8),
        axis.text.x = element_text(hjust = 1,vjust = 1,angle = 45),
        axis.text.y = element_text(angle = 90,hjust = 0.5,vjust = 0.5),
        legend.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold",size = 12),
        aspect.ratio = 1/2)
```

**IRSNB 3121**

```{r results='hide'}
# Carcharocles megalodon.
cooper_megladon <- 
  openxlsx::read.xlsx(xlsxFile = "data/Cooper et al. 2022.xlsx",sheet = 1) |> 
  rename("Diameter" = 9) |> 
  mutate(Diameter = Diameter/10) |> 
  mutate(log_cd = log10(Diameter)) |>  
  dplyr::select(log_cd)

# Predictions (using C. carcharias as the reference)
meg_fit1 <- 
  10^predict(object = length_linear_fit_ols,newdata = data.frame(log_cd = cooper_megladon$log_cd),interval = "prediction") |> 
  as.data.frame() |>
  ggplot(aes(x = 1:141,y = fit)) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), color = "#014240") +
  geom_point(color = "#014240", fill = "white", shape = 21) +
  scale_y_continuous(limits = c(100,1200)) +
  geom_hline(yintercept = c(920,1110), linewidth = 1.2,
             color = c("#CA0020","#5D4B3C")) +
  labs(x = "Vertebrae (IRSNB 3121)", y = "Total length (TL, cm)",
       title = "Otodus megalodon",
       subtitle = "C. carcharias as the reference species",
       caption = "") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold.italic"),
        axis.title = element_text(face = "bold"),
        axis.text.y = element_text(angle = 90,hjust = 1,vjust = 1),
        aspect.ratio = 1)

round(100 - (1021.1817/100)/11.1 * 100,digits = 2) # 8% difference between regression-model estimate and minimum column length of 11.1 m.

# Predictions (using Lamnna nasus as the reference)
meg_fit2 <-
  10^predict(object = slope_intercept$model[[9]],newdata = data.frame(log_cd = cooper_megladon$log_cd),interval = "prediction") |> 
  as.data.frame() |> 
  ggplot(aes(x = 1:141,y = fit)) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), color = "#014240") +
  geom_point(color = "#014240", fill = "white", shape = 21) +
  scale_y_continuous(limits = c(100,1200)) +
  geom_hline(yintercept = c(1110), linewidth = 1.2,
             color = c("#5D4B3C")) +
  labs(x = "Vertebrae (IRSNB 3121)", y = "Total length (TL, cm)",
       title = "Otodus megalodon",
       subtitle = "Lamna nasus as the reference species",
       caption = "") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold.italic"),
        axis.title = element_text(face = "bold"),
        axis.text.y = element_text(angle = 90,hjust = 1,vjust = 1),
        aspect.ratio = 1)

# Predictions (using interspecific model)
meg_fit3 <-
  10^predict(object = interSM3,newdata = data.frame(log_cd = cooper_megladon$log_cd),interval = "prediction") |> 
  as.data.frame() |> 
  ggplot(aes(x = 1:141,y = fit)) +
  geom_errorbar(aes(ymin = lwr, ymax = upr), color = "#08306b") +
  geom_point(color = "#08306b", fill = "white", shape = 21) +
  scale_y_continuous(limits = c(100,1200)) +
  geom_hline(yintercept = c(1110), linewidth = 1.2,
             color = c("#B2182B")) +
  labs(x = "Vertebrae (IRSNB 3121)", y = "Total length (TL, cm)",
       title = "Otodus megalodon",
       subtitle = "",
       caption = "") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold.italic"),
        axis.title = element_text(face = "bold"),
        axis.text.y = element_text(angle = 90,hjust = 1,vjust = 1),
        aspect.ratio = 1)
```

## Body size evolution

```{r results='hide'}
shimada <- 
  openxlsx::read.xlsx(xlsxFile = "data/Shimada et al. 2021.xlsx") |> 
  mutate(TL = as.numeric(TL),
         Min = as.numeric(Min),
         Max = as.numeric(Max))

bzE <- ggplot(shimada) + 
  geom_segment(aes(x = Min, xend = Max, y = Genus, yend = Genus), linewidth = 1) +
  # geom_point(aes(x = (Min + Max) / 2, y = Genus, size = TL), color = "red") 
  geom_point(aes(x = (Min + Max) / 2, y = Genus, size = TL, color = TL >= 500), show.legend = T) +
  scale_color_manual(values = c("#006B81", "#B83A4B")) +
  labs(title = "Lamniform body size evolution", subtitle = "Estimated maximum possible total body length", x = "Geological Age (Ma)", y = NULL) +
  # geom_vline(xintercept = c(149.2,145.0,139.8,132.6,125.77,121.4,113.0,100.5,93.9,89.8,86.3,83.6,72.1,66.0,61.6,59.2,56.0,47.8,41.2,37.71,33.9,
  #                           27.82,23.03,20.44,15.98,13.82,11.63,7.246,5.333,3.600,2.58,1.80,0.774,0.129,0.0117,0.0), col = "black") +
  # geom_vline(xintercept = c(145.0,100.5,66.0,56.0,33.9,23.03,5.333,2.58,0.0117), col = "red") +
  geom_vline(xintercept = 66.0, col = "red") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold"),
        axis.text.y = element_text(face = "italic"),
        axis.title = element_text(face = "bold"))

bzEe <- shimada |> 
  filter(TL >= 450) |> 
  ggplot() + 
  geom_segment(aes(x = Min, xend = Max, y = Genus, yend = Genus), linewidth = 1) +
  geom_point(aes(x = (Min + Max) / 2, y = Genus, size = TL), color = "red") +
  labs(title = "Lamniform body size evolution", subtitle = "Estimated maximum possible total body length", x = "Geological Age (Ma)", y = NULL) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold"),
        axis.text = element_text(face = "bold"),
        axis.title = element_text(face = "bold"))
```

## Statement on data use

The data used in this study was compiled from multiple sources and future use, whether in part or whole, should cite and credit the original work.

## Download data

{{< downloadthis data/sharkData.xlsx dname="Lamniform shark data" label="Download lamniform data used in this study as .xlsx" icon="file-earmark-spreadsheet" type="success" >}}

## References

```{r}

```

